---
title: "Genomic data ecosystems compared"
subtitle: "dartRverse and the genometranslator–radr suite"
author: "Thierry Gosselin"
date: "2026-09-02"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  freeze: auto
---

## Purpose and scope

This report compares the dartRverse suite (including `dartR.base` version
1.2.3, dated 2026-02-17) with the ecosystem built around
`genometranslator`, `radr`, and their companion packages. It is a technical
audit of two different designs, not a benchmark and not a claim that one
package suite is universally superior.

The distinction matters because the comparison is broader than two filtering
packages. dartRverse centres its workflow on the `genlight` object and retains
strong links to DArT genotype reports and DArT-derived locus metrics. The
`genometranslator`–`radr` ecosystem centres its workflow on persistent GDS,
standardized genomic evidence, recalculation from the active data, and explicit
filter provenance. These choices lead to different strengths, limitations, and
levels of readiness for population-scale and whole-genome sequencing data.

The comparison asks four practical questions:

1. Which input formats and biological quantities are actually retained?
2. Are statistics recomputed from the active samples and markers?
3. Can filtering decisions be reconstructed and audited later?
4. What information can legitimately be written to downstream formats?

The distinction between phylogenetic interoperability and population-genomic
quality control is important. dartR has long had broad phylogenetic and
specialist export coverage, and its population-genomic functionality has grown
substantially over the last 6 years. 

The `radr` ecosystem prioritizes raw genotype evidence, population-aware
quality control, scalable repeated filtering, and provenance.

## Why researchers should control genotyping and assembly

This report is not an argument against DArT genotyping products. I have used
them extensively, and they can be extremely valuable for marker discovery,
exploratory analyses, pilot studies, and projects in which access to raw reads
or computational expertise is limited. The concern is narrower: for a detailed
population-genomic investigation, researchers should be able to understand,
test, and reproduce the steps that convert their reads into genotypes.

Much of my work concerns species with low population differentiation. When the
biological signal is small, it may be comparable in magnitude to technical
variation introduced by library, plate, sequencing run, assembly, genotype
calling, missing-data thresholds, or allele-frequency filters. Small changes
in pipeline parameters can alter the retained loci, site-frequency spectrum,
population structure, and downstream estimates. Empirical comparisons have
shown that RADseq assembly, mapping, genotyping, and filtering choices can
materially change population-genetic inference (Andrews et al., 2016; Shafer
et al., 2017).

For that reason, I need to know what happened to the reads. At minimum, a
researcher should be able to answer:

- How were reads demultiplexed, trimmed, and screened for quality?
- Was locus construction de novo or reference guided, and which parameters
  controlled locus assembly and matching among individuals?
- Which aligner, reference assembly, and mapping-quality rules were used?
- Which genotype caller and model generated the calls?
- Which depth, likelihood, quality, allele-balance, missingness, and
  allele-frequency thresholds were applied?
- Were those statistics recalculated after samples or markers were removed?
- Were library, plate, lane, project, or sequencing-run effects examined?
- Can the complete workflow be rerun from raw reads with recorded software
  versions, commands, parameters, and random seeds?

This is not a demand that genotypes be called *manually*. On the contrary, the
workflow should be scripted, versioned, and automated. Researchers do not need
to invent an assembler or genotype caller, but they should control the choice
of software and parameters and retain the intermediate evidence needed to
evaluate the result.

The practical barrier is now much lower than it was when reduced-representation
sequencing first became common. Detailed reviews, protocols, maintained
manuals, forums, reproducible examples, institutional support, and genomics
workshops are widely available. Stacks, for example, provides an integrated
and tunable workflow for de novo and reference-guided RADseq processing
(Catchen et al., 2013; Rochette & Catchen, 2017; Rochette et al., 2019), with
an extensive [online manual](https://catchenlab.life.illinois.edu/stacks/manual/).
Training a student to run, document, and interrogate such a workflow is a
valuable part of the scientific analysis rather than an optional technical
exercise.

The objective is not control for its own sake. It is to distinguish a weak
biological signal from a plausible technical signal, perform sensitivity
analyses, and make the path from reads to inference inspectable by collaborators,
reviewers, and future users of the data.

## Overview of current genomic formats

File formats should be evaluated by what they preserve, not only by whether a
program can read them. A format may be excellent for a specific analysis while
being unsuitable as the primary record of sequencing evidence.

| Format or family | Principal strength | Information normally retained | Recommended role |
|---|---|---|---|
| **VCF/BCF** | Standardized variant interchange | Coordinates, REF/ALT alleles, multiallelic records, filters, INFO annotations, `GT`, phase, and extensible FORMAT fields such as `GL`/`PL`, `AD`, `DP`, and `GQ` | **Preferred source and archival exchange format.** Prefer `GT` with `GL` or `PL`; retain allele depth and quality fields when available |
| **SeqArray GDS** | Compressed, indexed, disk-backed analytical storage | VCF-like variants and annotations in arrays that support selection, random access, and chunked computation | **Preferred persistent working format** for `genometranslator` and `radr` after import |
| **PLINK 2 PGEN/PVAR/PSAM** | Very fast, compact genotype analysis | REF/ALT tracking, hard calls, dosages, phase, multiallelic variants, selected VCF-style metadata, phenotypes, and covariates | Excellent **derived analysis format** for PLINK 2 workflows; generally not the sole archival record of sequencing evidence |
| **PLINK 1 BED/BIM/FAM** | Broad compatibility and efficient biallelic hard-call storage | Compact genotype dosage/calls, basic variant coordinates and alleles, sample/family fields | Useful derived compatibility format; substantially less expressive than VCF or PGEN |
| **DArT two-row allele counts** | Retains allele-specific DArT read support | Per-individual counts for both alleles, from which depth, balance, and genotypes can be evaluated | Preferred DArT input when VCF is unavailable; convert to standardized GDS for analysis |
| **DArT one- or two-row genotype reports** | Convenient vendor genotype delivery | Called genotypes and vendor locus summaries | Usable, but less informative than allele counts because the evidence behind each call is reduced or absent |
| **Genepop, FSTAT, and STRUCTURE-style genotype matrices** | Simple exchange with established population-genetics programs | Primarily sample, population, locus, and genotype codes | Legacy analysis/exchange formats only; do not use as the authoritative sequencing-data archive |

### VCF/BCF, GDS, and PLINK serve different purposes

- **VCF or BCF is the evidence-preserving interchange format.** The VCF
  specification defines variant alleles, coordinates, filters, annotations,
  genotype calls, genotype likelihoods, depth, quality, and phase (Danecek et
  al., 2011). BCF is its binary representation. This is the best starting point
  when the caller can provide `GT` with `GL` or `PL`, and preferably `AD`, `DP`,
  `GQ`, and genotype filters. See the official
  [VCF 4.5 specification](https://samtools.github.io/hts-specs/VCFv4.5.pdf)
  and [HTS specifications repository](https://github.com/samtools/hts-specs).
- **GDS is the persistent analytical representation.** SeqArray reorganizes
  VCF-like information into compressed arrays that can be selected and read in
  chunks without materializing the complete genotype matrix (Zheng et al.,
  2017). In this ecosystem, VCF/BCF is imported once and GDS becomes the
  filtered, auditable working object.
- **PLINK is an excellent derived analysis representation.** PLINK 2's
  PGEN/PVAR/PSAM files improve on BED/BIM/FAM by tracking REF/ALT alleles and
  supporting dosage, phase, multiallelic variants, and additional metadata.
  They are optimized for PLINK operations and broad analytical compatibility
  (Purcell et al., 2007; Chang et al., 2015). See the official
  [PLINK 2 format reference](https://www.cog-genomics.org/plink/2.0/formats).
- **Conversion is directional.** A rich VCF can be converted to a purpose-built
  PLINK dataset. If likelihoods, allele depths, filters, or annotations are
  discarded during that conversion, converting back to VCF cannot recreate
  them. The original VCF/BCF and the GDS audit trail should therefore be kept.

For this ecosystem the recommended sequence is:

1. Preserve the original VCF/BCF, preferably with `GT`, `GL` or `PL`, and the
   available depth and quality fields.
2. Import it with `genometranslator::read_genome()` to create or open the
   persistent GDS working representation.
3. Perform QC and filtering from the active GDS state with complete audits.
4. Export PGEN or BED only when a downstream PLINK-compatible analysis needs
   it.

### Why legacy population-genetics formats are insufficient

Genepop, FSTAT, and STRUCTURE-style formats were designed for multilocus
genotype analyses rather than for preserving sequencing evidence. They remain
useful as narrow inputs to the programs and methods for which they were
created (Goudet, 1995; Pritchard et al., 2000; Rousset, 2008). The problem is
using them as the master copy of RADseq or whole-genome data.

For RADseq, conversion to a legacy genotype matrix commonly loses or obscures:

- per-sample REF and ALT read depth, genotype likelihood, and genotype quality;
- the relationship among multiple SNPs originating from the same read or RAD
  locus;
- read position, locus sequence, reference placement, and technical metrics;
- site and genotype filters, multiallelic state, phase, and explicit missingness
  provenance; and
- enough information to recalculate depth- and confidence-based statistics
  after samples are removed.

These losses matter because RADseq QC is not only a question of whether a cell
contains a genotype. Allele dropout, paralogy, uneven depth, batch structure,
linked SNPs on the same sequence, and genotype uncertainty can all produce a
plausible-looking genotype matrix.

At whole-genome scale, the limitations become more severe:

- plain-text genotype matrices grow very large and are slow to parse repeatedly;
- most operations require reading or materializing the full matrix rather than
  selecting compressed genomic ranges or annotation fields;
- variant coordinates and alleles may be represented minimally or by local
  conventions rather than a sequencing-variant schema;
- rich per-variant and per-genotype annotations have no standard destination;
  and
- programs developed for relatively small sets of approximately independent
  loci may require aggressive LD pruning and model-specific preparation before
  millions of linked variants can be analysed responsibly.

The last point separates the **file format** from the **analysis model**. For
example, the original STRUCTURE model assumes loci are unlinked and in linkage
equilibrium within populations (Pritchard et al., 2000). Exporting a whole
genome to STRUCTURE text does not make all of its linked variants appropriate
for that model. A legacy export should be a deliberate, filtered endpoint—not
the beginning of quality control and not the archival dataset.

## The fundamental difference: the data model

The most important difference between the two ecosystems is not the number of
functions. It is the object used internally to represent genomic data.

dartR uses the `genlight` class from `adegenet` as its central data object.
The `genlight` class was introduced as a memory-efficient representation of
genome-wide SNP data. It stores, for each individual and locus, the number or
frequency of a reference allele using bit-level compression. This is extremely
efficient for biallelic SNP matrices and can accommodate different ploidies.
The biological matrix is complemented by locus and individual information,
which dartR commonly stores in the `@other` slot (Jombart & Ahmed, 2011).

The packages developed around `radr` use two complementary representations:

1. **SeqArray GDS is the persistent genomic data store.** GDS is a hierarchical,
   array-oriented, compressed, disk-backed format. SeqArray builds a
   sequencing-variant schema on GDS and supports efficient random access to
   samples, variants, genotypes, alleles, and VCF-like annotations without
   loading the entire dataset into memory (Zheng et al., 2017).
2. **Tibbles are the analytical and reporting representation.** Selected data
   are extracted from GDS into rectangular tables and manipulated with focused
   tidyverse packages. Under the tidy-data model, each variable is a column,
   each observation is a row, and each cell contains one value. Each type of
   observational unit belongs in its own table (Wickham, 2014; Wickham et al.,
   2019). For example, one table describes markers, another samples, and a long
   genotype table can use one row per sample-marker observation.

GDS and tibbles are therefore not alternatives. The GDS is the compressed
source of truth; tibbles are explicit, readable views used for calculations,
joins, summaries, audits, and figures. Results can be written back to GDS as
metadata or active selections without making the full genotype matrix an
in-memory table.

### What kinds of genomic data can each representation hold?

| Data type or property | `genlight` | SeqArray GDS used by `genometranslator` | Tibbles used by the ecosystem |
|---|---|---|---|
| Biallelic SNP genotypes | Native strength: compact allele dosage per individual and locus | Native and compact | Straightforward in either wide or long form |
| Variable ploidy | Supported by `genlight`; values represent allele counts or frequencies within the declared ploidy | Ploidy-aware genotype arrays can be stored when represented by the importer and schema | Can represent any ploidy explicitly, but conventions must be defined |
| Multiallelic variants | Not naturally represented as one dosage value at one locus; records generally need to be removed, reduced, or split | Native VCF-style variant records may contain multiple alternate alleles | Naturally represented using allele identifiers or separate rows, although potentially large |
| Phased genotypes and haplotypes | No native phase-aware haplotype representation in the core dosage matrix | Phased genotype information and related FORMAT annotations can be retained when available in the source | Can represent haplotypes explicitly in columns/rows, but without storage compression |
| Indels and complex alleles | A biallelic dosage could be stored, but many `genlight`/dartR paths assume single-base SNP allele labels | REF and ALT strings and associated annotations can represent SNPs, indels, and other VCF records | Can store arbitrary allele strings and variant classes |
| Structural-variant annotations | Not a native strength | Can store variant records and annotations hierarchically; support by a particular analysis remains function-specific | Can describe intervals, breakpoints, and annotations explicitly |
| Dominant presence/absence markers | Values can be stored, and dartR uses `genlight` for SilicoDArT, but biological meaning depends on external datatype conventions | Can be stored with an explicit marker type and separate presence/depth semantics | Naturally represented as logical or 0/1 observations |
| Per-sample total depth | Not part of the core `genlight` dosage matrix; may be supplied in additional structures or summarized metrics | Retained as a sample-by-variant FORMAT field when imported | Easily analysed as one variable in a long table |
| Per-sample REF and ALT depth | Not represented by the single dosage value | Retained as allele-depth arrays for compatible inputs | Easily represented, but expensive for whole genomes if fully materialized |
| Genotype likelihood or quality | Not part of the core dosage representation | Can retain VCF FORMAT annotations such as genotype quality or likelihood fields when imported | Can represent the values explicitly after selective extraction |
| Marker and sample metadata | Stored through `genlight` slots and dartR conventions such as `@other$loc.metrics` and `@other$ind.metrics` | Stored as GDS annotations with standardized extraction functions | Natural rectangular representation for joining, checking, and reporting metadata |
| Complete filter state | Successive objects and `@other$history` record much of the workflow | Active sample/variant selections and filter annotations persist in the GDS | Kept/removed audit tables make decisions directly inspectable |
| Whole-genome scale | Memory-efficient bit-level storage for biallelic dosage. dartR offers optional, documented-as-testing file-backed matrices, but functions that call operations such as `as.matrix(x)` still materialize complete or selected matrices in RAM | Disk-backed compressed storage designed for whole-genome variant data, with sample/variant selection, random access, and chunked processing without loading the complete dataset | Best used for selected subsets and summaries rather than the complete whole-genome genotype matrix |

The ability of a container to hold a data type does not imply that every
analysis can use it. For example, SeqArray can retain multiallelic variants and
complex annotations, while an HWE or LD function may deliberately require
filtered biallelic SNPs. The advantage is that unsupported information need not
be silently destroyed during import; it can be retained, inspected, and
filtered explicitly.

### Scale illustration: 2,000 individuals and 100,000 SNPs

Consider a modest whole-genome subset containing 2,000 individuals and 100,000
SNPs. It contains 200 million sample-by-variant genotype cells.

| Representation | Approximate implication |
|---|---|
| Ordinary R numeric matrix | 200 million double-precision values require about 1.6 GB before object overhead, metadata, missing-value structures, and temporary copies |
| Ordinary R integer matrix | About 800 MB for the genotype values alone, again before overhead and copies |
| Bit-packed biallelic dosage | The theoretical core genotype information is on the order of tens of megabytes; for example, two bits per diploid genotype would require about 50 MB, before object metadata and implementation overhead |
| Plain VCF containing only `GT` | Approximately 0.8 GB is plausible from three genotype characters plus a separator per cell, before site columns and headers; the exact size depends on genotype spelling and missingness |
| Plain VCF containing `GT:AD:DP:GQ` | Commonly several gigabytes because every cell also carries text representations of allele depth, total depth, and genotype quality |
| Compressed VCF/BCF or GDS | Size is highly data-dependent because missingness, allele frequencies, repeated strings, annotations, compression method, and depth fields all matter; dimensions alone cannot provide a defensible file-size prediction |

These values illustrate scale; they are not performance benchmarks. In
particular, comparing only file sizes can be misleading. A small `genlight`
object is valuable for analyses that require only biallelic dosage, whereas a
larger GDS may intentionally retain allele depths, quality fields, filters, and
annotations that the `genlight` object does not contain.

Opening a SeqArray GDS normally creates a connection and reads its structure
and selection metadata; it does not load all 200 million genotype cells.
Opening and closing can therefore take a fraction of a second to a few seconds
on a local disk, largely independently of the full genotype matrix. This is an
expectation, not a guaranteed timing: storage hardware, network filesystems,
GDS structure, validation checks, and active-filter restoration all contribute.
Importing the original VCF into GDS is a separate, one-time operation and can
take substantially longer.

Actual project-specific values should be reported rather than estimated. The
following pattern records file sizes and connection times without extracting
the complete genotype matrix:

```{r}
#| eval: false

vcf <- "study.vcf.gz"
gds.file <- "study.gds"

data.frame(
  file = c(vcf, gds.file),
  size_gib = file.info(c(vcf, gds.file))$size / 1024^3
)

open.time <- system.time({
  genome <- genometranslator::read_genome(
    data = gds.file,
    verbose = FALSE
  )
})

close.time <- system.time({
  SeqArray::seqClose(genome)
})

open.time
close.time
```

This measures the cost of reopening an already imported GDS. It should not be
presented as the time required to read or analyse every genotype. Extraction
time must be benchmarked separately for the particular subset and fields used
by an analysis.

### Why GDS was chosen

The decision to build around GDS was made approximately a decade ago, before
whole-genome resequencing became routine in many of these projects. It was a
strategic, long-term choice: reduced-representation datasets were already
growing, and it was clear that repeatedly materializing complete genomic
matrices in memory would not remain sustainable as projects moved toward whole
genomes.

That choice now allows the same overall workflow to span DArT, RADseq, and
whole-genome data while retaining richer evidence than a dosage matrix alone.
The objective was not merely to make today's filtering run faster. It was to
avoid redesigning the ecosystem when variant counts reached millions, depth
and allele balance became essential QC variables, and analyses needed to be
resumed from a persistent filtered state.

### Division of responsibilities around the data model

| Responsibility | Package |
|---|---|
| STACKS/genotyping workflows | `stackr` |
| Standardization | `standart` |
| Reference placement and genome-coordinate work | `scaffoldr` |
| Format detection, translation, metadata normalization, and GDS construction | `genometranslator` |
| Quality control, filtering, diagnostics, and genomic discovery | `radr` |
| Downstream individual and sex assignment | `assigner` |

This separation is intentional. `genometranslator` should not decide which
markers are biologically acceptable, and `radr` should not require a different
implementation for every input format. Once information is normalized in GDS,
the same QC method can operate on DArT, VCF, PLINK, or another supported source.

## Reading: format names are not enough

The preferred input is a **VCF containing called genotypes (`GT`) and genotype
likelihoods (`GL`) or Phred-scaled genotype likelihoods (`PL`)**. When the
sequencing pipeline provides them, allele depth (`AD`), total depth (`DP`),
genotype quality (`GQ`), and genotype-level filters (`FT`) should also be
retained. A called genotype alone records the pipeline's final decision;
likelihoods and read depths retain evidence needed to evaluate that decision.

The following table lists genotype inputs, rather than merely naming broad
format families. Preferred inputs are listed first. It reflects the inspected
package sources as of the date of this report.

| Preference and input | dartR reader | `genometranslator::read_genome()` detection and underlying reader | Important consequence |
|---|---|---|---|
| **1. VCF (`.vcf` or `.vcf.gz`) with `GT` and `GL` or `PL`** | `dartR.base::gl.read.vcf()` uses `vcfR::read.vcfR()` and converts calls to a `genlight` dosage matrix | Detects the VCF header or extension and dispatches to `genometranslator::read_vcf()` | `genometranslator` imports to SeqArray GDS and can retain the richer VCF record and FORMAT evidence. The inspected dartR conversion does not preserve a general sample-by-variant FORMAT model in `genlight` |
| **2. DArT two-row allele-count CSV** | No count-preserving reader was identified. `gl.read.dart()` recognizes one- and two-row genotype encodings, but its detector is not a reader for arbitrary per-allele counts | Detects DArT `counts` and dispatches to `genometranslator::read_dart()` | Preferred when DArT input is required: observed read support is retained separately for both alleles in every individual-marker cell |
| DArT two-row SNP genotype CSV | `dartR.base::gl.read.dart()`; automatically distinguishes the supported one- and two-row genotype reports | Detects DArT `2rows` and dispatches to `genometranslator::read_dart()` | Both read the called genotypes, but this is not equivalent to retaining per-allele counts |
| DArT one-row SNP genotype CSV | `dartR.base::gl.read.dart()` | Detects DArT `1row` and dispatches to `genometranslator::read_dart()` | Both supported; less underlying evidence is available than in the count report |
| SilicoDArT presence/absence CSV | `dartR.base::gl.read.silicodart()` | Detects SilicoDArT and dispatches to `genometranslator::read_dart()` | Both recognize dominant presence/absence markers |
| Count-valued SilicoDArT CSV | No distinct count-preserving route identified | Detects count-valued SilicoDArT through `genometranslator::read_dart()` | Presence/absence and its read support can remain distinct in GDS |
| PLINK binary files (`.bed`, `.bim`, `.fam`) | `dartR.base::gl.read.PLINK()` uses `snpStats::read.plink()` | Detects `.bed` plus companion files and dispatches to `genometranslator::read_plink()` | Both support binary PLINK, but convert it into different internal data models |
| PLINK text files (`.ped`, `.map`) | `dartR.base::gl.read.PLINK()` invokes external PLINK to make BED before conversion | Not the direct text route exposed by `read_genome()`; use a supported PLINK representation | dartR support depends on an external PLINK conversion |
| PLINK transposed files (`.tped`, `.tfam`) | No direct `gl.read.PLINK()` route documented in the inspected source | Detects `.tped` plus `.tfam` and dispatches to `genometranslator::read_plink()` | Directly supported by `genometranslator` |
| Generic SNP genotype CSV (`0/1/2` or nucleotide pairs) | `dartR.base::gl.read.csv()` | A standardized wide or long TSV/tibble is detected and normalized internally | The required table conventions differ; these are dosage/call tables, not substitutes for VCF likelihoods or allele depths |
| Aligned FASTA sequence files | `dartR.base::gl.read.fasta()`; biallelic SNPs are extracted and multiallelic sites skipped | No FASTA genotype reader is dispatched by `read_genome()` | A genuine dartR-only direct reader in this comparison |
| Genepop (`.gen`) | No direct core genotype reader identified in the inspected dartR exports | Detects and dispatches to `genometranslator::read_genepop()` | Normalized through the common interface |
| FSTAT (`.dat`) | No direct core genotype reader identified in the inspected dartR exports | Detects and dispatches to `genometranslator::read_fstat()` | Normalized through the common interface |
| SeqArray GDS (`.gds`) | No direct GDS-to-`genlight` reader among the inspected `gl.read.*()` functions | Detects a GDS file and opens it; an existing `SeqVarGDSClass` connection is also accepted | Persistent filters and annotations are restored without loading the complete genotype matrix |
| Apache Parquet (`.parquet`) | No direct genotype reader identified | Detects Parquet and normalizes its tabular content internally | Useful for package-native tabular data, but it is not a VCF-equivalent variant schema |
| Legacy RAD/FST (`.rad`) | No direct reader identified | Detects and reads the legacy package format internally; deprecated | Supported for migration, not recommended for new data |
| `genind`, `genlight`, or `gtypes` R objects, including `.rds` | `genlight` is dartR's native object | Detects the object/class and dispatches to `read_genind()`, `read_genlight()`, or `read_gtypes()` | Allows migration into the standardized ecosystem |

`dartR.popgen::gl.read.structure()` is not counted above as a genotype importer:
it reads output files produced by STRUCTURE and returns run summaries and
Q-matrices. It serves a useful but different purpose.

For `genometranslator`, users normally need only:

```{r}
#| eval: false

genome <- genometranslator::read_genome(
  data = "study.vcf.gz",
  strata = "study_strata.tsv"
)
```

`read_genome()` detects the supported input and dispatches to the appropriate
engine. The exported `read_*()` functions remain available when a user needs
format-specific arguments, but they are not required for routine import.

### Why VCF is preferred

VCF is preferred because it is a documented, community-maintained interchange
standard rather than a vendor-specific report layout. It defines variant
coordinates, REF and ALT alleles, filters, INFO annotations, and extensible
per-sample FORMAT fields (Danecek et al., 2011). In particular, `GT` records the genotype call, `GL`
records log10-scaled genotype likelihoods, and `PL` records the corresponding
Phred-scaled likelihoods. This makes the meanings and ordering of the values
portable between compliant tools. See the
[official VCF specification](https://samtools.github.io/hts-specs/VCFv4.5.pdf)
and the [HTS specifications repository](https://github.com/samtools/hts-specs).

The preferred VCF therefore contains `GT` plus `GL` or `PL`, rather than `GT`
alone. Retaining `AD`, `DP`, `GQ`, and `FT` where available further strengthens
quality control. It allows downstream analyses to revisit confidence, allele
balance, depth, and filters instead of treating every called genotype as
equally certain.

### If DArT input is required, prefer two-row allele counts

DArT deliverables are not interchangeable. The principal forms encountered by
this ecosystem are:

- **one-row SNP genotype reports**, containing a called diploid genotype per
  individual and marker;
- **two-row SNP genotype reports**, using separate binary rows for the two
  alleles but still representing genotype calls rather than read counts;
- **two-row allele-count reports**, containing the observed read support for
  each allele in each individual; and
- **SilicoDArT reports**, representing dominant presence/absence markers, which
  may also contain count-valued read support.

If a project must work from DArT output, the preferred input is the **two-row
allele-count report**, not merely a two-row binary genotype report. The two
rows preserve, for each individual and marker, the observed read depth
supporting each allele. `genometranslator::read_dart()` can retain those
allele-specific counts, total depth, and the derived genotype in GDS.

Allele-specific depth makes it possible to evaluate allelic balance in
heterozygotes, weakly supported calls, allele dropout, unusually deep markers,
batch effects, and minor allele depth. It also permits relevant summaries to be
recomputed after individuals are removed. A called genotype or an average
depth per marker cannot reconstruct this sample-by-marker evidence. The
inspected dartR import path reduces the data to a `genlight` genotype matrix
and marker-level metrics; it does not retain the full REF/ALT count matrix.

### What dartR actually does with DArT depth

The inspected `gl.read.dart()` documentation correctly states that both
one-row and two-row genotype reports are recognized. The internal detector,
however, uses the maximum sample value to decide whether the report contains
one genotype row or two binary allele rows. A genuine allele-count report can
contain values much larger than two and therefore falls outside that logic.

For ordinary DArT genotype input, dartR may calculate a marker-level `rdepth`
using the number of called individuals, genotype-derived allele-presence
proportions, and the imported DArT summary columns `AvgCountRef` and
`AvgCountSnp`. That quantity is an estimated average for the marker. It is not
the matrix of observed REF and ALT read counts for every sample-marker pair.

This distinction matters because two datasets may have the same genotype and
average marker depth but very different allelic balance, dropout, batch
effects, or support for rare alleles. For population-genomic QC, the count
report is therefore the preferred DArT input whenever available.

## Recalculation after samples are removed

It would also be incorrect to say that dartR never recalculates statistics.
`gl.read.dart(recalc = TRUE)` recalculates several genotype-derived locus
metrics, including call rate, observed genotype frequencies, and average PIC.
Many dartR filters maintain flags indicating whether a metric is current.

The limitation is narrower:

- DArT-supplied technical summaries that cannot be reconstructed from the
  `genlight` genotype matrix remain properties of the original DArT report.
- Some filters expose a `recalc` option and do not always enable it by default.
- Marker-level reconstructed depth cannot replace sample-by-marker REF and ALT
  counts that were never retained.
- A function call in `@other$history` records intent, but it is not a complete
  table of exactly which markers and samples were removed and retained.

In the GDS workflow, statistics are calculated from the currently active
samples and markers. When sample removal changes the dataset, downstream call
rate, allele count, allele frequency, monomorphism, and depth summaries are
regenerated from that active state. Filters write explicit kept and removed
tables and preserve a removal reason in marker/sample metadata.

## Detection and filtering

| QC question | dartR | radr | Material difference |
|---|---|---|---|
| Missing genotypes per marker | `gl.filter.callrate()` | `filter_genotyping()` | radr recomputes active-sample missingness and writes complete marker audits |
| Missingness and heterozygosity per individual | Reporting, subsetting, and call-rate tools | `filter_individuals()` | radr integrates sample missingness, heterozygosity, coverage, audit, and subsequent monomorphism checks |
| All-missing markers | `gl.filter.allna()` | `detect_all_missing()` plus relevant filters | radr separates diagnosis from removal |
| Rare alleles | `gl.filter.maf()` | `filter_ma(ma.stats = "mac" | "maf" | "mad")` | radr supports MAC, MAF, and minor allele depth; MAC is the preferred general-purpose criterion, while MAD uses actual allele-read support when available |
| REF/ALT calibration | Genlight dosage orientation and locus metrics | Count-, depth-, or ancestral-state calibration | radr can redefine the observed minor allele after sample filtering and distinguish count-based from depth-based evidence |
| Markers shared among strata | No close dedicated filter identified | `filter_common_markers()` | radr can require marker representation across defined strata |
| Monomorphic markers | `gl.filter.monomorphs()` | `filter_monomorphic()` and automatic integration into filters that alter the active samples | radr explicitly rechecks monomorphism after individual removal, when polymorphic status can change |
| DArT reproducibility | `gl.filter.reproducibility()` | `filter_dart_reproducibility()` | Both necessarily use a DArT pipeline metric; radr labels and audits that provenance |
| Read depth | `gl.filter.rdepth()` using available locus metrics | `filter_coverage()` using retained sample/marker depth and allele depth | Marker-average filtering and sample-by-marker evidence answer different QC questions |
| Secondary SNPs on reads | `gl.filter.secondaries()` | `filter_snp_number()` and `filter_snp_position_read()` | radr separates number-per-locus and position-on-read decisions |
| HWE | Exact or chi-square tests, by population or pooled; multiple-testing adjustment is optional and off by default | Fast exact mid-p testing, population thresholds, multiple threshold outputs, and integration with other QC evidence | dartR is not statistically trivial, but radr is more conservative operationally and provides richer sensitivity diagnostics |
| Excess heterozygosity | `gl.filter.excess.het()` tests HWE-like excess, initially focusing on loci with observed heterozygosity above 50% | `detect_paralogs()`, `detect_het_outliers()`, HWE, allele-balance, and coverage-aware diagnostics | Excess heterozygosity can flag candidates but is not, by itself, a reliable paralog diagnosis |
| Duplicate samples | Replicate and Hamming-distance functions | `detect_duplicate_genomes()` | radr produces an explicit diagnostic/audit workflow |
| Mixed or contaminated samples | No close dedicated equivalent found | `detect_mixed_genomes()` | radr-specific diagnostic |
| Identity by missingness | No close dedicated equivalent found | `detect_ibm()` | radr can expose wet-lab plate/lane structure encoded in missingness |
| Rare-allele support and biallelic problems | No close unified diagnostics | `detect_rare_alleles()`, `detect_biallelic_problems()` | radr separates weak read support for rare alleles from genotype-representation errors |
| Linkage disequilibrium | `gl.filter.ld()` | `filter_ld()` | radr operates on active genomic coordinates and records removed/retained markers |
| Sex-linked markers | `dartR.sexlinked` reports, retains, drops, and infers sex | `sexy_markers()` combines presence, heterozygosity, and normalized depth while treating SilicoDArT as dominant | radr preserves evidence by method and produces an assignment-ready panel without interpreting dominant data as diploid heterozygosity |
| Inversion candidates | No equivalent identified in the inspected modules | `detect_inversions()` | radr-specific discovery workflow |

### Hardy--Weinberg filtering: a fairer characterization

The dartR implementation is more capable than a "naive HWE filter" label
suggests. It offers exact and chi-square tests, population-wise testing, a
threshold for the number of populations showing departure, and optional
multiple-testing correction. Its important operational defaults are a minimum
population size of five and no multiple-comparison adjustment.

The radr distinction is therefore not simply possession of a better exact
test. It is the surrounding workflow: exact mid-p calculations, conservative
default thresholds, diagnostics over several significance and population
thresholds, warnings about population structure and missingness, integration
with coverage and mixed-genome checks, and complete kept/removed outputs.

### Putative paralogs

dartR does address excess heterozygosity, so it should receive credit for
screening one common paralog signal. The biological caution is that excess
heterozygosity is nonspecific: duplication, collapsed loci, sex linkage,
population mixture, contamination, and genotyping error can produce similar
patterns. A defensible paralog assessment should combine genotype proportions
with coverage, allele balance, reproducibility, population consistency, and
sequence/reference evidence where these are available.

## Writing and the `gl2*` functions

dartR has impressive breadth of specialist exporters, particularly for
phylogenetics and established external programs. Breadth does not guarantee
lossless round-tripping, however. Every writer can export only the information
retained by the central object.

| Writer or pattern | What the inspected code does | Limitation or risk |
|---|---|---|
| `gl2vcf()` | Calls `gl2plink()`, creates PED/MAP intermediates, then invokes an external PLINK 1.9 binary to recode them as VCF | This is a PLINK-generated genotype VCF, not reconstruction of the source VCF. Original FORMAT fields, allele depths, genotype likelihoods, phasing, and most INFO annotations are not recoverable |
| DArT SNP to VCF | Can produce a minimal VCF if the `genlight` object has usable allele labels; unmapped coordinates are set to chromosome/position zero | So it can write *a* VCF from DArT SNP genotypes, contrary to a categorical claim that it cannot. It cannot write the original DArT allele counts or technical evidence into a faithful VCF |
| SilicoDArT to VCF/PLINK | `gl2plink()` does not visibly reject dominant data before converting values 0/1/2 to diploid allele pairs | Presence/absence value 1 risks being represented as a heterozygous diploid genotype; such conversion is biologically inappropriate without an explicit dominant-marker model |
| `gl2plink()` missing data | The inspected PED writer converts missing genotypes to `"0/0"`, then removes `/`, yielding the required PLINK missing allele pair `0 0` | The output is syntactically conventional, but the code is easy to misread as a reference homozygote; explicit tests are warranted |
| `gl2plink()` alleles | Extracts alleles using fixed character positions in strings such as `A/T` | Assumes simple single-character biallelic alleles and is unsuitable for indels or more complex allele strings |
| `gl2vcf()` coordinates | If the maximum current position is below 1000 and no genome-position field is supplied, positions become zero; chromosomes may also become zero | Read-relative SNP positions are deliberately discarded, but valid small genomic coordinates could be misclassified by the heuristic |
| `gl2vcf()` external command | Builds one shell command as a string and expects `plink` at `file.path(plink.bin.path, "plink")` | Paths containing spaces or shell-sensitive characters are fragile; exit status is not handled as explicitly as with `system2()` argument vectors |
| Temporary files | Uses fixed names such as `gl_plink_temp` in the output folder and `mylist.txt` in the session temp directory | Concurrent calls can collide; intermediate files can remain in the output directory |
| `gl2gds()` | Writes an SNPRelate GDS containing genotype, sample ID, variant ID, chromosome, position, and allele | It is not equivalent to the richer SeqArray GDS used by genometranslator: DArT metrics, sample metadata, per-genotype depth, allele counts, and filter provenance are not carried into the created GDS |
| Specialist `gl2*` exporters | Many correctly target PHYLIP, PAUP, SNAPPER, TreeMix, BayesAss, GenAlEx, STRUCTURE, and related programs | These are genuine dartR strengths, but many are terminal text exports rather than reversible translations |
| Return values | Several writers return `NULL`; some return a generated table invisibly | Programmatic composition and validation are inconsistent across writers |

### Why `gl2vcf()` is necessarily lossy

A `genlight` matrix principally stores genotype dosage and a limited set of
locus/sample annotations. Once a VCF has been reduced to that representation,
the following cannot generally be reconstructed:

- per-allele read depth (`AD`), total depth (`DP`), and genotype quality (`GQ`);
- genotype likelihoods (`GL`/`PL`) and phasing;
- multiallelic alleles removed during import;
- heterogeneous INFO fields and flags not retained in locus metrics;
- the complete original VCF header and provenance.

The name `gl2vcf()` is therefore accurate as an export operation, but its
output should not be interpreted as a lossless VCF round trip. A GDS generated
directly from the source VCF or DArT allele-count report retains substantially
more evidence for later filtering and re-export.

## Provenance and reproducibility

| Provenance level | dartR | genometranslator + radr |
|---|---|---|
| Function history | Calls are appended to `@other$history` by many functions | Function calls and arguments are written and filter parameters are retained |
| Current metric flags | `@other$loc.metrics.flags` tracks whether several metrics are believed current | Statistics are derived from the active GDS state when needed |
| Removed markers | Usually inferred by comparing objects or rerunning history | Explicit per-step removed-marker table with reason |
| Retained markers | Present in returned object | Explicit per-step retained-marker table plus active GDS selection |
| Removed/retained samples | Usually inferred from successive objects | Explicit sample audits for functions capable of sample removal |
| Reopening a project | Requires the correct serialized object and interpretation of its history | GDS reopens with sample/variant selection and filter metadata |
| Genotype evidence | Primarily dosage matrix plus selected metrics | Dosage, missingness, total depth, allele depth, and source metadata when available |

## Overall interpretation

dartR remains particularly strong as a broad `genlight` analysis and export
ecosystem. Its many phylogenetic and specialist writers are real advantages,
and recent versions contain considerably more population-genomic functionality
than older versions. The source audit also corrects three overly broad
criticisms: dartR does read ordinary two-row DArT genotype files, it does
recalculate several genotype-derived metrics, and `gl2vcf()` can generate a
minimal VCF from DArT SNP genotypes.

The central limitation for modern population-genomic QC is information loss.
The inspected dartR DArT reader does not retain two-row per-individual allele
counts; its VCF reader simplifies VCF into `genlight`; and many downstream
statistics can consequently use only genotype dosage or imported/reconstructed
marker summaries. Once raw evidence is discarded, later functions cannot
recreate it.

The `genometranslator -> radr -> assigner` design is strongest when the goal is
to maintain a traceable chain from raw genotype evidence to filtered markers
and downstream assignment:

1. ingest heterogeneous sources through one interface;
2. retain technical evidence in a disk-backed GDS;
3. recompute statistics from active samples and markers;
4. separate detection from removal;
5. audit every removal and retention decision; and
6. export only representations that are biologically valid for the marker type.

## Recommendations for the radr ecosystem

1. Continue treating the DArT allele-count report as the preferred DArT SNP
   input and explain why it differs from the two-row binary genotype report.
2. Record the engine used by every genometranslator adapter in its help page and
   run log, just as `vcfR` and `snpStats` should be visible in a dartR audit.
3. Maintain an explicit matrix of which fields survive every conversion.
4. Refuse or strongly warn about biologically invalid dominant-to-diploid
   conversions, particularly SilicoDArT to VCF, PLINK, or HWE workflows.
5. Recalculate monomorphism and allele statistics whenever active individuals
   change.
6. Keep assembly/genotyping responsibilities in `stackr`, normalization in
   `standart`, reference placement in `scaffoldr`, format translation in
   `genometranslator`, QC/discovery in `radr`, and assignment in `assigner`.
7. Add round-trip tests that compare genotypes, alleles, coordinates, missing
   values, metadata, and depth—not merely whether an output file was created.

## Audited source locations

The principal dartR files inspected for this report were:

- `dartR.base/R/gl.read.dart.r`
- `dartR.base/R/utils.read.dart.r`
- `dartR.base/R/utils.dart2genlight.r`
- `dartR.base/R/gl.read.vcf.r`
- `dartR.base/R/gl.read.PLINK.r`
- `dartR.base/R/gl2vcf.r`
- `dartR.base/R/gl2plink.r`
- `dartR.base/R/gl2gds.r`
- `dartR.base/R/gl.filter.callrate.r`
- `dartR.base/R/gl.filter.maf.r`
- `dartR.base/R/gl.filter.hwe.r`
- `dartR.base/R/gl.filter.excess.het.r`
- `dartR.sexlinked/R/gl.report.sexlinked.r`
- `dartR.sexlinked/R/gl.filter.sexlinked.r`
- `dartR.sexlinked/R/gl.infer.sex.r`

The principal local comparison files were:

- `genometranslator/R/read_genome.R`
- `genometranslator/R/dart.R`
- `genometranslator/R/vcf.R`
- `genometranslator/R/plink.R`
- `genometranslator/R/gds.R`
- `radr/R/filter_genotyping.R`
- `radr/R/filter_individuals.R`
- `radr/R/filter_ma.R`
- `radr/R/filter_monomorphic.R`
- `radr/R/filter_common_markers.R`
- `radr/R/filter_hwe.R`
- `radr/R/filter_coverage.R`
- `radr/R/detect_paralogs.R`
- `radr/R/sex_markers.R`

Because this is a source-snapshot audit, findings should be rechecked when any
of the compared implementations changes.

## References

Andrews, K. R., Good, J. M., Miller, M. R., Luikart, G., & Hohenlohe, P. A.
(2016). Harnessing the power of RADseq for ecological and evolutionary
genomics. *Nature Reviews Genetics*, *17*, 81--92.
https://doi.org/10.1038/nrg.2015.28

Catchen, J., Hohenlohe, P. A., Bassham, S., Amores, A., & Cresko, W. A. (2013).
Stacks: An analysis tool set for population genomics. *Molecular Ecology*,
*22*(11), 3124--3140. https://doi.org/10.1111/mec.12354

Chang, C. C., Chow, C. C., Tellier, L. C. A. M., Vattikuti, S., Purcell, S. M.,
& Lee, J. J. (2015). Second-generation PLINK: Rising to the challenge of larger
and richer datasets. *GigaScience*, *4*, 7.
https://doi.org/10.1186/s13742-015-0047-8

Danecek, P., Auton, A., Abecasis, G., Albers, C. A., Banks, E., DePristo,
M. A., Handsaker, R. E., Lunter, G., Marth, G. T., Sherry, S. T., McVean, G.,
& Durbin, R. (2011). The variant call format and VCFtools. *Bioinformatics*,
*27*(15), 2156--2158. https://doi.org/10.1093/bioinformatics/btr330. Current
format specification: https://samtools.github.io/hts-specs/VCFv4.5.pdf

Graffelman, J., & Moreno, V. (2013). The mid p-value in exact tests for
Hardy--Weinberg equilibrium. *Statistical Applications in Genetics and
Molecular Biology*, *12*(4), 433--448.

Gruber, B., Unmack, P. J., Berry, O. F., & Georges, A. (2018). dartr: An R
package to facilitate analysis of SNP data generated from reduced
representation genome sequencing. *Molecular Ecology Resources*, *18*(3),
691--699. https://doi.org/10.1111/1755-0998.12745

Goudet, J. (1995). FSTAT (version 1.2): A computer program to calculate
F-statistics. *Journal of Heredity*, *86*(6), 485--486.
https://doi.org/10.1093/oxfordjournals.jhered.a111627

Jombart, T., & Ahmed, I. (2011). adegenet 1.3-1: New tools for the analysis of
genome-wide SNP data. *Bioinformatics*, *27*(21), 3070--3071.
https://doi.org/10.1093/bioinformatics/btr521

Linck, E., & Battey, C. J. (2019). Minor allele frequency thresholds strongly
affect population structure inference with genomic data sets. *Molecular
Ecology Resources*, *19*(3), 639--647.

Purcell, S., Neale, B., Todd-Brown, K., Thomas, L., Ferreira, M. A. R., Bender,
D., et al. (2007). PLINK: A tool set for whole-genome association and
population-based linkage analyses. *The American Journal of Human Genetics*,
*81*(3), 559--575.

Pritchard, J. K., Stephens, M., & Donnelly, P. (2000). Inference of population
structure using multilocus genotype data. *Genetics*, *155*(2), 945--959.

Rousset, F. (2008). genepop'007: A complete re-implementation of the Genepop
software for Windows and Linux. *Molecular Ecology Resources*, *8*(1),
103--106. https://doi.org/10.1111/j.1471-8286.2007.01931.x

Rochette, N. C., & Catchen, J. M. (2017). Deriving genotypes from RAD-seq
short-read data using Stacks. *Nature Protocols*, *12*, 2640--2659.
https://doi.org/10.1038/nprot.2017.123

Rochette, N. C., Rivera-Colon, A. G., & Catchen, J. M. (2019). Stacks 2:
Analytical methods for paired-end sequencing improve RADseq-based population
genomics. *Molecular Ecology*, *28*(21), 4737--4754.
https://doi.org/10.1111/mec.15253

Robledo-Ruiz, D. A., Austin, L., Amos, J. N., Castrejon-Figueroa, J., Harley,
D. K., Magrath, M. J. L., Sunnucks, P., & Pavlova, A. (2023). Easy-to-use R
functions to separate reduced-representation genomic datasets into sex-linked
and autosomal loci, and conduct sex assignment. *Molecular Ecology Resources*.

Shafer, A. B. A., Peart, C. R., Tusso, S., Maayan, I., Brelsford, A., Wheat,
C. W., & Wolf, J. B. W. (2017). Bioinformatic processing of RAD-seq data
dramatically impacts downstream population genetic inference. *Methods in
Ecology and Evolution*, *8*(8), 907--917.
https://doi.org/10.1111/2041-210X.12700

Waples, R. S. (2015). Testing for Hardy--Weinberg proportions: Have we lost the
plot? *Journal of Heredity*, *106*(1), 1--19.

Wigginton, J. E., Cutler, D. J., & Abecasis, G. R. (2005). A note on exact
tests of Hardy--Weinberg equilibrium. *The American Journal of Human Genetics*,
*76*(5), 887--893.

Wickham, H. (2014). Tidy data. *Journal of Statistical Software*, *59*(10),
1--23. https://doi.org/10.18637/jss.v059.i10

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., Francois, R.,
Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L.,
Miller, E., Bache, S. M., Muller, K., Ooms, J., Robinson, D., Seidel, D. P.,
Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K., & Yutani, H. (2019).
Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686.
https://doi.org/10.21105/joss.01686

Zheng, X., Gogarten, S. M., Lawrence, M., Stilp, A., Conomos, M. P., Weir,
B. S., Laurie, C., & Levine, D. (2017). SeqArray---a storage-efficient
high-performance data format for WGS variant calls. *Bioinformatics*, *33*(15),
2251--2257. https://doi.org/10.1093/bioinformatics/btx145
