---
title: "Detecting and validating candidate sex-linked markers with radr"
author: "Floriaan Devloo-Delva and Thierry Gosselin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting and validating candidate sex-linked markers with radr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope

Sex-linked markers can help identify the sex-determination system of a species,
describe sex-chromosome differentiation, prevent sex-linked variation from
confounding population-genetic analyses, and assign genetic sex when external
sex characters are absent or unreliable.

`radr::sexy_markers()` screens an already prepared GDS dataset for markers whose
presence, heterozygosity, or normalized read depth differs between recorded
females and males. It is designed for diploid, biallelic SNP datasets, including
RADseq, DArTseq, and related reduced-representation data. The function does not
alter the GDS. It respects the active sample and marker selections, restores
them before returning, and writes an auditable results folder.

<div class="alert alert-warning" role="alert">
<strong>A candidate is not proof of a sex-chromosome location.</strong>
Sex-specific dropout, plate or library effects, population structure, paralogy,
mapping bias, and errors in recorded sex can reproduce the expected patterns.
Discovery should be followed by technical, genomic, and biological validation.
</div>

## Biological background

In an XX/XY system, females are the homogametic sex and males are the
heterogametic sex. In a ZZ/ZW system, males are homogametic and females are
heterogametic. Sex chromosomes can range from nearly homomorphic chromosomes
with a small non-recombining region to strongly differentiated chromosomes with
extensive sequence loss and divergence. Consequently, no single marker pattern
is expected in every species.

Reduced-representation data can reveal three complementary signals.

### Sex-specific presence

A sequence confined to the heterogametic chromosome should be present mainly in
one sex:

* a Y-like marker is expected mainly in males;
* a W-like marker is expected mainly in females.

These markers are especially useful for genetic sex assignment. However, a
similar presence/absence pattern can be produced by low read depth,
sex-confounded batches, allelic dropout, or a polymorphic restriction site.

### Sex-biased heterozygosity

Markers in homologous or gametologous regions can exhibit different genotype
patterns between sexes. Under a simple model, an X-like marker may show greater
heterozygosity in females than males, whereas a Z-like marker may show the
reverse. The direction and magnitude depend on the divergence and mapping
behaviour of the chromosome copies; a biological sex-linked marker does not
have to satisfy this simple expectation.

### Sex-biased read depth

Copy-number differences can also produce depth differences. Under an idealized
model, an X-linked region can have greater normalized depth in XX females than
XY males, while a Z-linked region can have greater normalized depth in ZZ males
than ZW females. Library size, capture efficiency, paralogy, and mapping bias
can all affect this signal, so `sexy_markers()` normalizes depth within samples
and still treats the result as candidate evidence.

## Origin of the method

The original `sexy_markers` method was developed by Floriaan Devloo-Delva,
Thierry Gosselin, Robin B. Thomson, and collaborators and described by
Devloo-Delva et al. (2024). The study introduced a genotyped-data workflow based
on the three signals above and evaluated it with 558 White Sharks
(*Carcharodon carcharias*) and 23,393 SNPs. It recovered nine Y-linked and 406
X-linked candidates, supported sex assignment, and led to a PCR assay for
independent validation.

The study also showed why study design matters. In that White Shark dataset,
subsamples with fewer than 100 individuals, fewer than 10,000 markers or half
of the available markers, and unbalanced sex ratios produced more false
positives. These values are empirical results from one species and dataset, not
universal thresholds. They are nevertheless a useful warning that a small,
unbalanced discovery set should be interpreted conservatively.

The current `radr` implementation is a redesign rather than an exact copy of
the published `radiator` function. It uses GDS input only, reads markers in
chunks, applies explicit effect-size rules, calculates method-specific tests
and false-discovery rates, audits metadata for confounding with recorded sex,
and produces a separate Y/W panel for downstream assignment. Cite
Devloo-Delva et al. (2024) for the biological method and original validation,
and record the `radr` version used for every new analysis.

## Prepare the discovery dataset

Complete the ordinary sample and marker quality-control workflow before the
final screen, but do not automatically discard every marker with sex-biased
missingness: sex-specific absence may be the biological signal of interest.
Inspect missingness before and after filtering and retain the original audit
files.

At minimum:

1. verify individual identifiers and recorded sex;
2. remove failed samples, obvious duplicates, and unresolved mixtures;
3. inspect genotype call rate, depth, allele balance, and paralogy;
4. record plate, library, lane, sequencing run, population, collection site,
   project, and year when available;
5. check whether recorded sex is confounded with any technical or biological
   group; and
6. retain enough known females and males for discovery and independent
   validation.

The metadata must contain `INDIVIDUALS` and the column named by `sex.column`.
The values `F`, `female`, `M`, and `male` are recognized without regard to case.
Other values are treated as unknown and are not used to discover candidates.

```{r, eval=FALSE}
sample_metadata <- readr::read_tsv("sample_metadata.tsv")

table(sample_metadata$SEX, useNA = "ifany")
table(sample_metadata$SEX, sample_metadata$PLATE, useNA = "ifany")
table(sample_metadata$SEX, sample_metadata$POPULATION, useNA = "ifany")
```

If sex and population or batch are inseparable, the data cannot distinguish a
sex effect from that other effect. More modelling cannot recover information
that the sampling design does not contain. Add balanced samples or validate the
markers in independent plates, populations, or projects.

## Run the screen

```{r, eval=FALSE}
sex_markers <- radr::sexy_markers(
  data = "study.filtered.gds",
  strata = sample_metadata,
  sex.column = "SEX",
  presence.threshold = 0.90,
  absence.threshold = 0.10,
  min.heterozygosity.difference = 0.20,
  coverage.ratio.threshold = 1.50,
  coverage.threshold = 1,
  min.samples.per.sex = 5,
  fdr.threshold = 0.05,
  require.significance = TRUE,
  path.folder = getwd()
)

sex_markers
sex_markers$candidates
```

The permissive default `require.significance = FALSE` selects markers from the
effect-size rules alone. This is useful for exploration and for very small
datasets in which formal tests have low power. For a final discovery panel,
`require.significance = TRUE` is usually preferable: a marker must then satisfy
both the effect-size rule and the FDR threshold for its method. Significance is
not a substitute for a meaningful effect size or external validation.

### Thresholds

`presence.threshold` and `absence.threshold` define Y-like and W-like
presence/absence patterns. For example, `0.90` and `0.10` require the marker to
be detected in at least 90% of the expected sex and no more than 10% of the
other sex.

`min.heterozygosity.difference` is the minimum absolute difference between
female and male heterozygosity. `coverage.ratio.threshold` is the minimum ratio
between the larger and smaller mean normalized depths. Thresholds should be
specified before examining the candidate identities when possible, then tested
in sensitivity analyses.

`coverage.threshold` defines read presence when depth is available. If retained
depth is unavailable, the function uses a non-missing genotype as presence and
does not run the normalized-depth method.

### Statistical tests

Presence and heterozygosity are tested with two-sample score tests for
proportions. Normalized depth is tested with Welch tests on
`log2(1 + normalized depth)`. Benjamini-Hochberg FDR correction is applied
separately to each method because the methods test different observations and
hypotheses.

Within each sample, read depth is divided by its mean positive depth over the
active markers. This reduces sample-wide sequencing-depth differences. It does
not remove marker-by-batch interactions or confounding between sex and plate,
lane, library, population, or collection group.

## Understand the result

The returned `sexy_markers` object contains:

```{r, eval=FALSE}
names(sex_markers)

sex_markers$statistics
sex_markers$candidates
sex_markers$assignment_panel
sex_markers$sample_summary
sex_markers$metadata_audit
sex_markers$plots
```

The direction of every statistic is female minus male:

| Quantity | Positive value | Negative value |
|---|---|---|
| `presence_difference` | more frequent in females | more frequent in males |
| `heterozygosity_difference` | more heterozygous in females | more heterozygous in males |
| `log2(coverage_ratio_female_male)` | greater normalized depth in females | greater normalized depth in males |

Therefore, a strong negative presence difference is Y-like and a strong
positive difference is W-like. Positive heterozygosity or depth evidence is
labelled X-like, and negative evidence is labelled Z-like. These are operational
labels for expected patterns, not direct chromosome annotations.

A marker can be selected by more than one method. Agreement among methods is
useful evidence, but disagreement is not automatically an error: young,
homomorphic, diverged, repetitive, or incompletely assembled sex chromosomes
can generate different combinations of signals.

## Files written

Each run creates a dated results folder containing:

| File | Purpose |
|---|---|
| `radr_sexy_markers_args_*.tsv` | complete argument record |
| `sex_marker_statistics.tsv` | statistics and candidate flags for every tested marker |
| `candidate_sex_markers.tsv` | markers passing at least one candidate rule |
| `sex_assignment_panel.tsv` | assignment-ready Y-like and W-like markers |
| `sex_sample_summary.tsv` | numbers and mean positive depth by recorded sex |
| `sex_metadata_audit.tsv` | association of recorded sex with metadata columns |
| `candidate_sex_markers.fasta` | candidate sequences, when sequence metadata exist |
| `sex_markers_presence.*` | presence effect versus FDR plot |
| `sex_markers_heterozygosity.*` | heterozygosity effect versus FDR plot |
| `sex_markers_coverage.*` | normalized-depth effect versus FDR plot, when available |

The metadata audit reports Cramer's V between recorded sex and each available
categorical metadata variable. A strong association is a warning about possible
confounding; it is not proof that the candidate is technical, and it is not an
adjusted association test. Review the underlying contingency table and the
experimental design.

## Validate candidates

Discovery and validation answer different questions. Reproducing the recorded
sex of the same individuals used to discover a panel measures resubstitution,
not out-of-sample accuracy.

A strong validation plan includes several levels:

1. **Technical validation:** inspect per-sample genotypes, depth, allele balance,
   missingness, and sequence uniqueness. Repeat libraries when possible.
2. **Design validation:** test the panel in different plates, lanes, projects,
   sampling years, and laboratories.
3. **Population validation:** test independent populations because sex
   chromosomes and restriction-site polymorphisms can vary geographically.
4. **Genomic validation:** map sequences to a chromosome-scale assembly and
   test whether candidates co-localize in a plausible sex-linked region.
5. **Biological validation:** genotype independently sexed animals and report
   sensitivity, specificity, ambiguous calls, and confidence intervals.
6. **Assay validation:** develop and test an independent PCR, amplicon, or SNP
   assay for the final operational marker panel.

Use a holdout set or cross-validation during panel development, but retain a
fully independent final validation set whenever the intended application is
management, monitoring, or clinical-quality sex assignment.

## Pass the panel to assigner

`radr` discovers and audits candidate markers. Genetic sex assignment belongs
in `assigner`, which keeps prediction separate from discovery and does not
overwrite recorded sex.

```{r, eval=FALSE}
assignment <- assigner::assign_genetic_sex(
  data = "study.filtered.gds",
  panel = sex_markers$assignment_panel,
  strata = sample_metadata,
  sex.column = "SEX",
  path.folder = getwd()
)

assignment$assignments
assignment$comparison
```

The assignment panel intentionally contains Y-like and W-like presence markers,
which provide direct directional evidence for the heterogametic sex. X-like and
Z-like candidates remain valuable for biological interpretation and custom
models, but they are not silently converted into presence-based assignment
rules.

## Alternative methods

No method is best for every input type or sex-chromosome system. Applying more
than one method can be informative when their assumptions and input data are
understood.

| Method | Suitable input and main signal | When it is useful |
|---|---|---|
| **RADSex** | Demultiplexed RADseq reads; sex-biased marker presence and depth distributions | A fast, purpose-built workflow starting from reads, without requiring an existing SNP GDS |
| **dartR / dartR.sexlinked** | Reduced-representation SNPs in `genlight`; sex-linked genotype and call-rate patterns | An R workflow for identifying, filtering, and using multiple classes of sex-linked loci |
| **SDpop** | Genotypes or RNA-seq from natural populations; hierarchical models of sex-linkage and gametologs | Model-based inference of XY or ZW systems and posterior evidence for sex-linked genes, especially with mapped genomic data |
| **OutFLANK** | Genotyped SNPs partitioned by sex; allele-frequency differentiation | A generic outlier scan that can reveal differentiated homologous regions but is not a sex-marker-specific model |
| **pcadapt** | Genotyped SNPs; association with principal components | A complementary structure-based outlier scan, but results can reflect population or technical structure unrelated to sex |

RADSex begins closer to raw RADseq reads and is especially attractive when a
SNP dataset has not already been produced. `dartR.sexlinked` provides a broader
genlight-based workflow that can identify sex-linked classes and infer sex.
SDpop is a probabilistic population-genomic method with different data and model
requirements. OutFLANK and pcadapt are generic outlier methods: they can detect
allele-frequency differences between recorded sexes but do not by themselves
establish sex linkage.

### Why the current radr workflow differs from dartR.sexlinked

The comparison below is based on `dartR.sexlinked` 1.2.2, source commit
`4bb887a` dated 23 March 2026. Software changes, so the package version should
always accompany a methodological comparison. Floriaan Devloo-Delva is listed
as an author of `dartR.sexlinked`; the present comparison concerns the current
code and inferential workflow, not contributor credit.

`dartR.sexlinked` is useful when data already live in the dartR `genlight`
ecosystem, the investigator is prepared to declare an XY or ZW system in
advance, and identifying several genotype-based classes for filtering or
assignment is the main objective. Its published companion workflow was tested
in birds and a mammal and is an important alternative to `radr`.

For a new `radr` analysis, however, `radr::sexy_markers()` has several deliberate
advantages:

* **The hypotheses are not forced by a declared system.** `radr` reports Y-,
  W-, X-, and Z-like patterns from their observed directions. The current
  dartR functions require `system = "xy"` or `system = "zw"` before screening.
* **Effect sizes are explicit and configurable.** `radr` requires user-recorded
  presence, absence, heterozygosity-difference, and depth-ratio thresholds. In
  the reviewed dartR code, Y/W detection uses a fixed call rate of 0.10 in the
  expected absent sex and FDR of 0.01, without also requiring a high call rate
  in the expected present sex. X/Z and gametolog labels require FDR at most
  0.01 and the expected direction, but no minimum heterozygosity difference.
* **Observed zero counts remain zero.** The reviewed dartR implementation
  changes every zero cell to one before its Fisher or chi-square tests. This
  prevents numerical failures but changes the observed contingency table,
  effect estimate, and p-value. `radr` does not add this pseudocount.
* **Read depth is an independent evidence stream.** `radr` can use retained GDS
  depth, normalized within samples, to detect X/Z-like copy-number differences.
  The reviewed dartR screen uses genotype call rate and heterozygosity but does
  not implement the read-depth scenario from Devloo-Delva et al. (2024).
* **Technical confounding is reported.** `radr` audits the association between
  recorded sex and plate, lane, library, population, project, and other supplied
  metadata. The reviewed dartR workflow does not perform an equivalent audit or
  adjust its marker tests for these variables.
* **Large datasets do not require one full R genotype matrix.** `radr` reads the
  active GDS selection in chunks and restores it unchanged. The reviewed dartR
  implementation converts the complete `genlight` object to a conventional R
  matrix before analysis, which can become memory intensive.
* **Discovery is separated from assignment.** `radr` writes a versioned,
  marker-level Y/W panel. `assigner::assign_genetic_sex()` applies explicit
  agreement and margin thresholds, retains recorded sex for comparison, and
  returns `U` when evidence is tied or insufficient. The reviewed
  `dartR.sexlinked::gl.infer.sex()` combines presence calls with two K-means
  classifications; its result depends on a random seed unless one is supplied,
  and disagreement is returned as a starred majority assignment for manual
  review rather than an explicit unresolved class. In the reviewed code, equal
  support defaults to the first ordered label, and three missing preliminary
  assignments can consequently produce `*F` instead of an unresolved result.
* **The complete analysis is auditable.** `radr` writes the arguments, all marker
  statistics, candidate table, assignment panel, sample summary, metadata
  audit, sequences, and diagnostics to a dated folder. It also has automated
  tests for GDS selection restoration, classification direction, metadata
  handling, and ambiguous assignment. The reviewed `dartR.sexlinked` repository
  does not contain an automated test suite.

These are not cosmetic differences. Replacing zero counts, omitting minimum
effect sizes, or ignoring a sex-by-batch association can change which markers
are called. Conversely, `radr` is not automatically correct because it is more
auditable: its thresholds still require biological justification, and every
panel still requires independent validation.

Devloo-Delva et al. (2024) compared the original `sexy_markers` workflow with
`dartR::gl.report.sexlinked()`, OutFLANK, and pcadapt in the White Shark data.
The methods recovered markedly different candidate sets, which is expected
because they target different signals. Candidate count alone is therefore not
a measure of accuracy. Compare genomic localization, replication, assignment
performance, and independent assay validation.

## Reporting checklist

Report enough detail for the screen to be reproduced and challenged:

* sample counts by recorded sex, population, and technical batch;
* how sex was recorded and how uncertain records were treated;
* the starting and active numbers of samples, loci, and SNPs;
* all upstream filters and the complete marker audit;
* whether genotype calls or read depth defined marker presence;
* all effect-size, FDR, depth, and minimum-sample thresholds;
* the `radr` version and saved argument file;
* candidate counts by evidence class and overlap among classes;
* metadata associations and known design confounding;
* validation data that were not used for discovery; and
* sensitivity, specificity, ambiguous-call rate, and population transferability
  for any operational sex-assignment panel.

## References

Devloo-Delva F, Gosselin T, Butcher PA, Grewe PM, Huveneers C, Thomson RB,
Werry JM, Feutry P (2024) An R-based tool for identifying sex-linked markers
from restriction site-associated DNA sequencing with applications to
elasmobranch conservation. *Conservation Genetics Resources*, 16, 11-16.
[https://doi.org/10.1007/s12686-023-01331-5](https://doi.org/10.1007/s12686-023-01331-5)

Feron R, Pan Q, Wen M, Imarazene B, Jouanno E, Anderson J, Herpin A, Journot L,
Parrinello H, Klopp C, Kottler VA, Roco AS, Du K, Kneitz S, Adolfi M, Wilson CA,
McCluskey B, Amores A, Desvignes T, Goetz FW, Takanashi A, Kawaguchi M,
Detrich HW, Oliveira MA, Nóbrega RH, Sakamoto T, Nakamoto M, Wargelius A,
Karlsen Ø, Wang Z, Stöck M, Waterhouse RM, Braasch I, Postlethwait JH,
Schartl M, Guiguen Y (2021) RADSex: a computational workflow to study sex
determination using restriction site-associated DNA sequencing data.
*Molecular Ecology Resources*, 21, 1715-1731.
[https://doi.org/10.1111/1755-0998.13360](https://doi.org/10.1111/1755-0998.13360)

Gruber B, Unmack PJ, Berry OF, Georges A (2018) dartr: an R package to
facilitate analysis of SNP data generated from reduced representation genome
sequencing. *Molecular Ecology Resources*, 18, 691-699.
[https://doi.org/10.1111/1755-0998.12745](https://doi.org/10.1111/1755-0998.12745)

Käfer J, Lartillot N, Marais GAB, Picard F (2021) Detecting sex-linked genes
using genotyped individuals sampled in natural populations. *Genetics*, 218,
iyab053.
[https://doi.org/10.1093/genetics/iyab053](https://doi.org/10.1093/genetics/iyab053)

Luu K, Bazin E, Blum MGB (2017) pcadapt: an R package to perform genome scans
for selection based on principal component analysis. *Molecular Ecology
Resources*, 17, 67-77.
[https://doi.org/10.1111/1755-0998.12592](https://doi.org/10.1111/1755-0998.12592)

Mijangos JL, Gruber B, Berry O, Pacioni C, Georges A (2022) dartR v2: an
accessible genetic analysis platform for conservation, ecology and
agriculture. *Methods in Ecology and Evolution*, 13, 2150-2158.
[https://doi.org/10.1111/2041-210X.13918](https://doi.org/10.1111/2041-210X.13918)

Robledo-Ruiz DA, Austin L, Amos JN, Castrejón-Figueroa J, Harley DKP,
Magrath MJL, Sunnucks P, Pavlova A (2025) Easy-to-use R functions to separate
reduced-representation genomic datasets into sex-linked and autosomal loci,
and conduct sex assignment. *Molecular Ecology Resources*, 25, e13844.
[https://doi.org/10.1111/1755-0998.13844](https://doi.org/10.1111/1755-0998.13844)

Whitlock MC, Lotterhos KE (2015) Reliable detection of loci responsible for
local adaptation: inference of a null model through trimming the distribution
of FST. *The American Naturalist*, 186, S24-S36.
[https://doi.org/10.1086/682949](https://doi.org/10.1086/682949)
