---
title: "Detecting candidate chromosomal inversions with radr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting candidate chromosomal inversions with radr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope

The analytical starting points are Li and Ralph (2019) for local PCA, Faria et
al. (2019) and Wellenreuther and Bernatchez (2018) for inversion biology and
interpretation, and empirical work combining genotype groups, LD,
recombination, diversity, and structural evidence. These references should be
read before treating a local-structure signal as an inversion.

`detect_inversions()` is a screening tool for finding genomic regions whose
local population structure is unusual relative to the rest of a GDS dataset.
It is designed for diploid, biallelic SNP data, including carefully filtered
RADseq and similar reduced-representation datasets.

The function detects **candidate inversion-associated haploblocks**. SNP data
can show the expected population-genomic footprint of an inversion, but they do
not directly demonstrate reversed chromosome orientation or locate physical
breakpoints. Use long reads, split reads, discordant read pairs, linkage maps,
comparative assemblies, or cytogenetics for structural confirmation.

<div class="alert alert-warning" role="alert">
<strong>Caution: do not scan unfiltered or batch-confounded data.</strong>
Structured missingness, uneven depth, allele dropout, paralogs, close relatives,
and plate, lane, library, sequencing-run, or genotype-calling effects can create
local PCA clusters, heterozygosity differences, and extended LD that closely
resemble an inversion signal. Perform dataset-level quality control first, avoid
confounding biological groups with technical batches, and cross-tabulate every
candidate genotype group against the available technical metadata.
<code>detect_inversions()</code> is not a substitute for data filtering or batch
assessment.
</div>

<div class="alert alert-info" role="alert">
<strong>Screen before LD pruning.</strong> Extended LD is part of the signal
being investigated. Run the first inversion screen before LD pruning, then use
the inferred regions and arrangements in explicit sensitivity analyses.
</div>

## How `radr` builds on previous inversion work

`radr::detect_inversions()` builds on analytical principles explored by Li and
Ralph (2019), who formalised local PCA for detecting changes in population
structure along a genome; Huang et al. (2020), who demonstrated that candidate
inversions can be recovered from reduced-representation SNP data; and Pearse
et al. (2019) and Akopyan et al. (2025), who showed the value of combining
regional genotype groups, linkage disequilibrium, recombination, diversity,
and structural evidence when interpreting inversion-associated haploblocks.

The function is an independent implementation rather than a wrapper
around another inversion package. It was designed to make these transferable
ideas practical for the quality-controlled RADseq and similar datasets already
used in `radr`. In particular, `detect_inversions()`:

* reads genotypes directly from GDS;
* constructs every window within a chromosome, linkage group, or scaffold;
* offers fixed-SNP, fixed-base-pair, and experimental LD-scaled windows;
* combines local covariance PCA with contiguous-window candidate detection;
* evaluates regional PCA groups quantitatively instead of treating a requested
  three-group k-means solution as evidence by itself;
* summarises heterozygosity, overall and within-group LD, candidate boundaries,
  flanking windows, internal transitions, and window-size sensitivity;
* returns ordered putative arrangement genotypes, relative call confidence,
  arrangement-specific whitelists, and homokaryotype-only sensitivity sets;
* accepts centromere, low-recombination, repeat, and assembly-gap annotations
  without allowing them to determine the candidate calls;
* explicitly records the consequences of missing-data filtering and temporary
  mean imputation for RADseq interpretation; and
* writes reproducible tables and publication-ready `ggplot2` diagnostics to a
  standard results folder while returning all results for further exploration.

This integrated workflow is the main advantage of using `radr`: it moves from
a broad local-structure scan to an auditable candidate evidence table without
claiming more resolution than the marker data provide. The output is intended
to guide the next experiment, not replace it. Linkage mapping can test
recombination suppression; haplotagging or other linked-read approaches can
phase the alternative haplotypes and refine structural hypotheses; and ONT or
PacBio long reads, breakpoint PCR, comparative assemblies, or cytogenetics can
provide physical confirmation. Haplotagging has recovered long population
haplotypes and inversion-associated barcode-sharing patterns at scale (Meier
et al. 2021), but exact or repetitive breakpoints may still require continuous
long reads or a junction-specific assay.

## What is a chromosomal inversion?

An inversion occurs when a chromosome segment is reversed. Individuals may
carry two copies of one arrangement, one copy of each arrangement, or two
copies of the alternative arrangement. These are often described as two
homokaryotypes and one heterokaryotype.

Recombination is commonly reduced between alternative arrangements in
heterokaryotypes. Consequently, alleles across a large interval can remain
associated as a haplotype. This can protect combinations of locally adaptive
alleles from being broken apart by recombination, especially when populations
exchange migrants. Inversions can therefore maintain genomic differentiation
even when much of the genome shows weak population structure.

```{r inversion-concept, echo=FALSE, results="asis"}
inversion_figure <- knitr::image_uri("figures/chromosomal-inversion.png")
cat(
  '<img src="', inversion_figure,
  '" alt="Three-panel diagram showing a chromosome segment reversing ',
  'orientation, formation of an inversion loop in a heterokaryotype, and ',
  'the linkage disequilibrium, local PCA groups, and possible population ',
  'FST peak detectable from SNP data." style="width:100%;height:auto;">',
  sep = ""
)
```

Not every inversion is adaptive, and not every long differentiated haploblock
is an inversion. Selection, centromeres, low-recombination regions, recent
admixture, family structure, paralogs, assembly errors, marker-density changes,
and technical batches can generate partially similar patterns.

## Types of chromosomal inversion and how to distinguish them

The classical distinction depends on the relationship between the inverted
segment and the centromere, not on the size of the candidate or the number of
genotype clusters.

```{r inversion-types, echo=FALSE, results="asis"}
inversion_types_figure <- knitr::image_uri("figures/inversion-types.png")
cat(
  '<img src="', inversion_types_figure,
  '" alt="Three-panel diagram showing paracentric-compatible, ',
  'pericentric-compatible, and undetermined candidate inversion ',
  'classifications relative to the centromere." ',
  'style="width:100%;height:auto;">',
  sep = ""
)
```

The labels in this schematic deliberately describe **compatibility**, not a
structurally confirmed inversion type. Classification requires credible
breakpoints and an independently supported centromere position. A
threshold-defined candidate core should remain **undetermined** when either is
uncertain. For a more detailed illustration of inversion-loop pairing and
meiotic products, see the figure by Bérénice Bougas in Wellenreuther and
Bernatchez (2018),
[https://doi.org/10.1016/j.tree.2018.04.002](https://doi.org/10.1016/j.tree.2018.04.002).

| Inversion type | Structural definition | Classical meiotic consequence in a heterokaryotype | Evidence needed for classification |
|---|---|---|---|
| **Paracentric** | Both breakpoints occur on the same chromosome arm, so the centromere is outside the inverted segment. | A crossover within the inversion loop can produce dicentric and acentric recombinant chromatids. Their reduced recovery contributes to apparent recombination suppression. | Two validated breakpoints on the same annotated chromosome arm. |
| **Pericentric** | The breakpoints occur on opposite chromosome arms, so the inverted segment includes the centromere. | A crossover within the inversion loop can produce recombinant chromatids carrying duplications and deletions. Reduced recovery can suppress observed recombination. | Validated breakpoints on opposite sides of a reliably annotated centromere. |

These are descriptions of simple inversions. Real chromosomes can instead
contain nested, overlapping, adjacent, or serial inversions, and an inversion
may be accompanied by duplications, deletions, translocations, repetitive
sequence, or assembly differences. A broad population-genomic signal can
therefore represent one simple inversion, several linked rearrangements, or a
larger structurally complex haploblock.

### What evidence separates the types?

Classification normally requires all of the following:

1. **A chromosome-level reference framework.** The candidate must be placed on
   an assembled chromosome with credible chromosome-arm and centromere
   annotation. The largest observed SNP position is not sufficient.
2. **Breakpoint evidence.** Split reads, discordant read pairs, long reads,
   breakpoint-spanning assemblies, optical maps, or breakpoint PCR should
   identify the two orientation changes. Repetitive breakpoint regions may
   require long continuous reads or assembly comparison.
3. **Centromere context.** The validated breakpoints must then be compared with
   centromere coordinates. Breakpoints on one arm support a paracentric
   classification; breakpoints on opposite arms support a pericentric one.
4. **Independent consistency checks.** Linkage maps, pedigrees, cytogenetics,
   synteny with another assembly, and patterns of dosage or read depth can test
   the proposed structure and expose more complex alternatives.

Centromere annotations are uncertain in many non-model fish assemblies. Merely
overlapping a putative centromeric region, showing suppressed recombination, or
changing chromosome-arm proportions is supportive context rather than decisive
classification. The reference assembly and coordinate version should always be
reported.

Local PCA, three arrangement-like clusters, heterozygosity, LD, and windowed
population differentiation can identify an inversion-associated haploblock,
but they do not reveal chromosome orientation and cannot by themselves
distinguish a paracentric inversion from a pericentric inversion. Until both
breakpoints and the centromere relationship are established, report the result
as a candidate region or putative inversion rather than assigning an inversion
type.

## Evidence expected from SNP data

A convincing candidate normally combines several lines of evidence:

1. **Local structure:** contiguous windows show a population structure that
   differs from the genomic background.
2. **Three genotype groups:** regional PCA may show two homokaryotype groups
   and an intermediate heterokaryotype group.
3. **Heterozygosity:** the intermediate group is often more heterozygous within
   the region than the two outer groups.
4. **Extended LD:** SNPs across the region show elevated linkage disequilibrium.
5. **Genomic continuity:** the signal spans adjacent markers and does not merely
   follow an isolated outlier SNP.
6. **Technical independence:** the signal does not track plates, lanes,
   libraries, missingness, coverage, or other processing variables.
7. **Population differentiation:** when biologically defined populations differ
   in arrangement frequency, a broad windowed $F_{ST}$ peak overlaps the local
   structure signal and helps prioritise the region for follow-up.

None of these criteria alone proves an inversion. For example, LD may be
uneven inside an old inversion because gene conversion and double crossovers
permit some gene flux between arrangements.

## Lessons from empirical fish examples

Published fish studies illustrate both the value and the limits of an
inversion screen. `detect_inversions()` does not reproduce their complete
analyses: several relied on whole-genome sequence, linkage maps, pedigrees,
long reads, or phenotype data that are not present in a typical RADseq GDS.
Instead, the function uses the parts of their reasoning that transfer to
marker data and reports candidates for further investigation.

### Rainbow trout: a structurally validated double inversion

Pearse et al. (2019) characterized a roughly 55-Mb double-inversion supergene
on rainbow trout chromosome Omy05. Linkage mapping showed almost complete
recombination suppression in heterokaryotypic parents, while comparative
genomics and long-read assembly helped resolve the structure. The two
arrangements were associated with migratory tendency, with effects that
depended on sex and dominance.

This example motivates several choices in `detect_inversions()`:

* evidence should extend across a contiguous haploblock rather than a single
  strongly associated marker;
* regional genotype groups, LD, heterozygosity, and boundary changes should be
  interpreted together;
* one broad signal may contain adjacent or nested rearrangements and should not
  automatically be described as one simple inversion; and
* association with sex, migration, or another phenotype can support biological
  relevance, but it does not establish chromosome orientation.

The outer candidate coordinates returned by `detect_inversions()` therefore
describe a marker-supported interval. Internal changes in window scores, LD,
or clustering may justify examining more than one rearrangement inside that
interval.

### Chinook salmon: a major-effect locus is not necessarily an inversion

Thompson et al. (2020) showed that variation near `GREB1L` is strongly
associated with adult migration timing in Chinook salmon. This is an important
counterexample for interpreting a genomic peak: a locus with a large
phenotypic effect, strong differentiation, or extended haplotypes is not by
itself evidence of a chromosomal inversion.

Consequently, `detect_inversions()` does not use phenotype association or an
isolated differentiation peak as sufficient evidence. A narrow selected region
may be biologically important without producing the extended local-PCA,
three-group, heterozygosity, and LD pattern expected from an inversion
haploblock.

### Atlantic silverside: distinguish inversions from centromeres

Akopyan et al. (2025) compared 168 Atlantic silverside genomes from four
populations using whole-genome variation and recombination maps. Large,
abruptly bounded differentiation haploblocks coincided with known inversions
and showed the characteristic three tight regional-PCA clusters. Narrower
differentiation peaks frequently coincided with putative centromeres. Regional
PCA in centromeric regions could also reflect reduced recombination, but the
individuals were more dispersed and retained more haplotype variation than in
the inversion regions.

The study also demonstrates why relative differentiation alone can mislead.
Elevated $F_{ST}$ near centromeres resulted partly from low within-population
diversity, whereas absolute sequence divergence ($d_{XY}$) was elevated in the
large inversions but reduced near centromeres. For `radr`, the transferable
diagnostic lessons are to:

1. compare broad, abrupt and contiguous haploblocks with narrow or gradual
   peaks;
2. examine the compactness and separation of the three inferred genotype
   groups, not merely the existence of a regional PCA pattern;
3. compare candidates with known or putative centromeres, recombination maps,
   marker density, repeats, and assembly gaps; and
4. avoid calling a low-recombination region an inversion without independent
   structural or linkage evidence.

Sparse, ascertained RADseq markers generally do not support the same robust
windowed $d_{XY}$ analysis as whole-genome sequence. `detect_inversions()`
therefore does not manufacture an absolute-divergence statistic from
insufficient data. Where dense sequence data and defensible population groups
are available, $F_{ST}$, diversity, and $d_{XY}$ provide valuable downstream
validation alongside the `radr` candidate scan.

### Cross-reference candidates with chromosome-resolved $F_{ST}$

Population differentiation is most informative when it is plotted along each
linkage group rather than reduced to one genome-wide value. The complementary
`assigner::fst_WC84()` function calculates Weir and Cockerham's (1984)
$F_{ST}$ from independently defined populations and, when `CHROM` and numeric
`POS` metadata are available, returns:

* `fst.linkage.groups`, with one summary per chromosome or linkage group;
* `fst.windows`, with WC84 variance components combined in physical windows;
  and
* `fst.genome.plot`, a chromosome-concatenated Manhattan-style figure.

The physical window size should reflect marker density and the expected scale
of the candidate. For sparse RADseq data, a 25-kb window may contain too few
markers; larger windows and a sensitivity analysis across several window sizes
are usually more defensible. Use the same genome assembly, chromosome labels,
position metadata, sample filters, and marker filters in both analyses.

```{r fst-cross-reference, eval=FALSE}
library(ggplot2)

strata <- readr::read_tsv("strata.tsv", show_col_types = FALSE)

fst <- assigner::fst_WC84(
  data = genome,
  strata = strata,
  linkage.group = TRUE,
  window.size = 1e6,
  window.method = "fixed",
  filename = "inversion_fst_crosscheck"
)

# Whole-genome overview supplied by assigner
fst$fst.genome.plot

# Linkage-group view with radr candidate intervals shaded
candidate_intervals <- inv$candidates |>
  dplyr::transmute(
    CHROM = as.character(chromosome),
    xmin = start,
    xmax = end
  )

ggplot(fst$fst.windows, aes(WINDOW_MID, FST_WC84)) +
  geom_rect(
    data = candidate_intervals,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill = "#E69F00",
    alpha = 0.18
  ) +
  geom_line(colour = "grey45") +
  geom_point(aes(colour = OUTLIER), size = 1.4) +
  facet_wrap(~ CHROM, scales = "free_x") +
  scale_x_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = " Mb")
  ) +
  labs(
    x = "Genomic position",
    y = expression("Windowed WC84 " * F[ST]),
    colour = "Exploratory outlier"
  ) +
  theme_bw()
```

An overlapping, spatially coherent $F_{ST}$ peak strengthens the case that a
candidate haploblock contributes to differentiation among the sampled
populations. Its absence does not reject an inversion: both arrangements may
occur at similar frequencies among populations. Conversely, an $F_{ST}$ peak
can arise from local selection, a centromere, low recombination, reduced
diversity, marker ascertainment, or technical structure without an inversion.

Population strata must be specified independently of the inversion scan.
Calculating $F_{ST}$ between the regional PCA clusters inferred from the same
candidate genotypes is circular and should not be presented as independent
support. Where whole-genome sequence is available, compare relative
differentiation with $d_{XY}$, within-population diversity, recombination, and
structural evidence, as illustrated by Akopyan et al. (2025).

### Rainbow trout across populations: inversion signals are context-dependent

Campbell, Anderson, Garza, and Pearse (2021) compared independently landlocked
rainbow trout populations with an anadromous source population. All landlocked
populations showed an increased frequency of the large Omy05 inversion, while
three of four showed an increased frequency of the Omy20 inversion. Outside
these major rearrangements, however, the genomic response to similar selection
was much less parallel.

This work reinforces two principles used in `detect_inversions()`. First, a long
inversion-associated haplotype can represent standing adaptive variation that
changes frequency repeatedly across populations. Second, neither the strength
of a genomic signal nor its association with migration should be assumed to be
constant across the species range. Candidate clusters should therefore be
cross-tabulated against population and environment, and biological associations
should be tested rather than assigned from the candidate region alone.

## How the local analysis is organised

Windows are created **within** chromosome, linkage-group, or scaffold
boundaries. If LG1 ends with 40 unused SNPs and LG2 follows in the GDS, those
40 SNPs are never combined with the first 60 SNPs of LG2. This prevents an
artificial window spanning unrelated genomic regions.

The analysis is not one PCA per linkage group. Instead:

1. each linkage group is divided into SNP windows;
2. a separate local covariance PCA is calculated for every window;
3. each window is represented by its leading covariance axes;
4. distances between window representations are calculated;
5. multidimensional scaling (MDS) places similar windows together and unusual
   windows farther from the genomic background;
6. contiguous unusual windows are joined into candidate regions;
7. a regional PCA, clustering, heterozygosity summary, and LD summary are
   calculated for each candidate.

The window calculations are independent. `parallel.core` can distribute them
across separate R workers. Each worker opens its own read-only GDS connection;
workers do not share one SeqArray connection. A progress bar reports completed
windows in sequential and parallel mode. Because every worker constructs a
sample covariance matrix, more workers also require more memory. Start with two
or four workers rather than automatically using every available core.

When several linkage groups are scanned, their window summaries are compared
in the same MDS analysis. The returned table retains the linkage-group label,
so chromosome-position plots should normally be faceted by linkage group. A
regional PCA belongs to one candidate on one linkage group and therefore does
not itself require an LG facet.

As a complementary diagnostic, `chromosome.pca = TRUE` also calculates one
independent PCA for every chromosome or linkage group. These chromosome-wide
PCAs do not select candidates. After a candidate has been inferred, its AA, AB,
and BB labels are projected as colours onto every panel. This makes it easy to
ask whether the same individual grouping is localized to the candidate linkage
group or recurs throughout the genome.

## Run a scan from GDS

The examples are not evaluated because they refer to user files.

```{r scan, eval=FALSE}
library(radr)

genome <- genometranslator::read_genome("filtered_dataset.gds")

sample_metadata <- readr::read_tsv("maintained_sample_metadata.tsv")

# Stage 1: ordinary discovery scan with the defaults
screen <- detect_inversions(
  data = genome,
  strata = sample_metadata,
  parallel.core = 4
)

screen$candidates
screen$chromosome.lengths

# Stage 2: a more detailed genome-wide scan
inv <- detect_inversions(
  data = genome,
  strata = sample_metadata,
  window.snps = 100,
  step.snps = 100,
  sensitivity.window.snps = c(100, 250, 500, 1000),
  min.call.rate = 0.90,
  min.candidate.windows = 2,
  stability.replicates = 100,
  parallel.core = 4,
  chromosome.pca = TRUE,
  chromosome.pca.max.snps = 2000,
  return.ld = TRUE
)

inv
inv$candidates
inv$arrangement.genotypes
inv$homokaryotype.whitelist
inv$homokaryotype.all.candidates
inv$path.folder
```

The first call is the normal starting point. It scans the filtered genome before
LD pruning and reports candidate regions without requiring the user to choose a
large set of tuning values. The second call adds SNP-window sensitivity,
assignment stability, retained LD matrices, and more parallel workers. These
extra diagnostics are valuable, but they need not be paid for before the basic
signal has been located.

`strata` requires `INDIVIDUALS`. Its rows act as a sample whitelist and its
other columns remain descriptive metadata. Include useful variables such as
`STRATA`, sequencing batch, library, lane, plate, extraction method, caller,
sampling year, or other project-specific factors. The function compares these
variables with regional PC1 and inferred arrangements, but does not copy them
into or use them to modify the GDS.

Window-size sensitivity and SNP-resampling stability are optional because both
add computation. `sensitivity.window.snps = NULL` and
`stability.replicates = 0` skip them. A rapid first screen can use those
defaults; a focused rerun of promising chromosomes can then request both.
Chromosome-wide PCA also adds computation. Set `chromosome.pca = FALSE` for a
minimal screen, or reduce `chromosome.pca.max.snps` when chromosomes contain
very dense whole-genome data. SNPs are selected at evenly distributed marker
indices within each chromosome so the cap preserves chromosome-wide coverage.

Known centromeres, assembly gaps, or other low-recombination annotations can be
supplied without allowing those annotations to determine which windows are
selected:

```{r known-regions, eval=FALSE}
known_regions <- data.frame(
  chromosome = c("LG5", "LG14", "LG14"),
  start = c(22000000, 18000000, 41000000),
  end = c(26000000, 23000000, 42500000),
  type = c("putative_centromere", "low_recombination", "assembly_gap")
)

inv <- detect_inversions(
  data = genome,
  window.snps = 100,
  step.snps = 50,
  known.regions = known_regions
)
```

The `known_region_overlap` column lists overlapping annotation types. An
overlap is a warning for interpretation, not evidence for or against an
inversion, and it does not change candidate selection or the evidence score.

## Chromosome size and candidate extent

When the GDS retains the original VCF contig dictionary, chromosome lengths are
read automatically. Candidate tables then report `candidate_span_bp`,
`chromosome_length_bp`, `chromosome_fraction`, `chromosome_percent`, and the
left and right flanking lengths. The complete length provenance is returned in
`inv$chromosome.lengths` and written to `chromosome_length_context.tsv`.

If declared lengths are absent from the GDS, provide either a named vector or a
table with `CHROM` and `LENGTH` columns:

```{r chromosome-lengths, eval=FALSE}
lengths <- readr::read_tsv("chromosome_lengths.tsv")

inv <- detect_inversions(
  data = genome,
  chromosome.lengths = lengths
)
```

A reference FASTA path or its existing `.fai` index can instead be supplied
with `reference.genome`. The function reads the index rather than scanning the
complete FASTA:

```{r reference-lengths, eval=FALSE}
inv <- detect_inversions(
  data = genome,
  reference.genome = "genome.fasta"
)
```

Without declared sequence lengths, the largest observed marker position is
used only as a labelled fallback. It underestimates chromosome length and can
therefore overestimate the percentage occupied by a candidate.

To investigate a known chromosome-14 signal without letting other linkage
groups define the background:

```{r chromosome, eval=FALSE}
chr14 <- detect_inversions(
  data = genome,
  strata = sample_metadata,
  chromosome = "14",
  window.snps = 100,
  step.snps = 50,
  min.call.rate = 0.90,
  min.candidate.windows = 2
)
```

Once this focused call confirms the signal, compare independent primary window
definitions. These calls genuinely recall the candidate boundaries:

```{r chromosome-boundary-sensitivity, eval=FALSE}
chr14_50_25 <- detect_inversions(
  data = genome, strata = sample_metadata, chromosome = "14",
  window.snps = 50, step.snps = 25, parallel.core = 4
)

chr14_100_50 <- detect_inversions(
  data = genome, strata = sample_metadata, chromosome = "14",
  window.snps = 100, step.snps = 50, parallel.core = 4
)

chr14_250_125 <- detect_inversions(
  data = genome, strata = sample_metadata, chromosome = "14",
  window.snps = 250, step.snps = 125, parallel.core = 4
)
```

Compare candidate start, end, span, chromosome percentage, and reciprocal
overlap across these runs. The shared interval is a defensible stable candidate
core; the union describes boundary uncertainty. Do not select only the run that
produces the cleanest or widest interval.

`sensitivity.window.snps` is complementary. It summarizes PC1 variance and LD
at additional scales within one run, but it does not independently rerun
candidate selection at every size:

```{r chromosome-summary-sensitivity, eval=FALSE}
chr14_sensitivity <- detect_inversions(
  data = genome,
  strata = sample_metadata,
  chromosome = "14",
  window.snps = 100,
  step.snps = 50,
  sensitivity.window.snps = c(50, 100, 250, 500),
  stability.replicates = 100,
  parallel.core = 4
)
```

After boundary sensitivity, repeat the candidate analysis with stricter
call-rate thresholds and alternative sample sets, particularly after removing
close relatives or suspect sequencing batches. Only then should stable marker
boundaries be compared with recombination maps, structural-variant callers,
long reads, read-pair evidence, assemblies, or breakpoint assays.

Overlapping windows (`step.snps < window.snps`) can help localise transitions,
but adjacent results are then strongly dependent. Coordinates should not be
reported as precise breakpoints merely because a score changes between two
overlapping windows.

For physical windows instead of equal-SNP windows, supply `window.bp`. This is
particularly useful for checking whether a candidate is robust to variation in
RAD marker density:

```{r physical-windows, eval=FALSE}
chr14_mb <- detect_inversions(
  data = genome,
  chromosome = "14",
  window.bp = 1e6,
  step.bp = 5e5,
  min.window.snps = 20
)
```

Experimental LD-scaled windows adapt their number of markers to local adjacent
LD. They help assess whether a signal depends on an arbitrary fixed marker
count, but they do not estimate an inversion breakpoint or a formal LD block:

```{r ld-scaled-windows, eval=FALSE}
chr14_ld <- detect_inversions(
  data = genome,
  chromosome = "14",
  window.method = "ld",
  ld.window.threshold = 0.10,
  ld.window.min.snps = 50,
  ld.window.max.snps = 500
)
```

## Plot window evidence by linkage group

`detect_inversions()` automatically saves the standard window-score, MDS, LD,
call-rate, regional-PCA, heterozygosity, and candidate-LD figures. It also
returns the `ggplot2` objects in `inv$output.files$plots`, so they can be changed
without rerunning the genomic calculations.

```{r window-plot, eval=FALSE}
library(ggplot2)

ggplot(
  inv$windows,
  aes(x = (start + end) / 2, y = robust_score,
      colour = candidate_window)
) +
  geom_point() +
  geom_hline(
    yintercept = inv$settings$score.threshold,
    linetype = 2
  ) +
  facet_wrap(~ chromosome, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6,
                                                    suffix = " Mb")) +
  labs(
    x = "Window midpoint",
    y = "Robust local-structure score",
    colour = "Candidate"
  ) +
  theme_bw()
```

This is the appropriate LG-faceted view: window score versus genomic position.
The MDS coordinates can also be inspected, but faceting them by linkage group
answers a different question because MDS axes represent similarity among
windows rather than physical position.

```{r mds-plot, eval=FALSE}
ggplot(inv$windows, aes(MDS1, MDS2, colour = chromosome,
                        shape = candidate_window)) +
  geom_point(size = 2) +
  theme_bw()
```

## Inspect candidate genotype groups

Each candidate has a matching entry in `inv$diagnostics`. A regional PCA plot
shows individuals, not windows:

```{r regional-pca, eval=FALSE}
candidate1 <- inv$diagnostics[[1]]

ggplot(candidate1$scores, aes(PC1, PC2, colour = arrangement)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(colour = "Putative arrangement genotype") +
  theme_bw()

candidate1$cluster_summary
candidate1$cluster_models
candidate1$assignment_stability
candidate1$metadata_audit
candidate1$metadata_contingency
candidate1$scores |>
  dplyr::select(individual, arrangement, arrangement_dosage,
                arrangement_confidence, assignment_stability)
```

When sample metadata are supplied, each candidate also writes a
`*_metadata_by_arrangement.tsv` table and one regional-PCA figure per usable
metadata variable. These outputs show directly whether inferred arrangements
are concentrated in particular populations, projects, plates, libraries, or
lanes. A strong association produces a console warning. This is a confounder
audit, not a correction: `strata` does not residualize or otherwise adjust the
local PCA. Where biological groups and wet-laboratory batches are confounded,
repeat the scan in balanced subsets or within independently replicated groups
before interpreting the candidate as an inversion.

## Compare candidate groups across linkage groups

The chromosome-PCA overview uses the arrangement calls from one candidate only
as colours. It does not rerun clustering or force three groups independently on
every linkage group. Every panel is calculated from genotypes on that linkage
group, using at most `chromosome.pca.max.snps` evenly distributed SNPs.

```{r chromosome-pca, eval=FALSE}
inv$chromosome.pca$summary
inv$output.files$plots$`INV-CAND-1_chromosome_pca`
```

Separation that is strongest on the candidate linkage group and weak elsewhere
supports a localized haploblock interpretation. Similar separation across many
linkage groups instead suggests genome-wide population structure, family
structure, admixture, or a technical batch effect that needs to be resolved
before describing the region as a putative inversion.

The reverse is not a formal rejection test. A short candidate interval can be
diluted by thousands of unrelated SNPs in a whole-linkage-group PCA, so weak or
absent separation in the candidate linkage-group panel does not invalidate a
clear regional PCA. Interpret this overview together with the window scan,
regional PCA, heterozygosity, LD, missingness, depth, sample metadata, and
window-size sensitivity.

For candidates with quantitative three-cluster support, `AA`, `AB`, and `BB`
mean the low, middle, and high regional-PC1 clusters. They are convenient
putative arrangement genotypes, not validated breakpoint genotypes. `AA` and
`BB` do not identify which arrangement is reference, ancestral, derived, or
physically inverted. Reversing the arbitrary sign of PC1 can exchange the two
outer labels without changing the biological result.

Candidates without quantitative three-cluster support use neutral `Group 1`,
`Group 2`, and `Group 3` labels. Their numeric cluster IDs remain in exported
tables, but arrangement dosage and homokaryotype classification are withheld.
Relative confidence describes separation from the next algorithmic cluster; it
is not a posterior probability. For supported candidates, the result folder
contains one whitelist per arrangement, one homokaryotype-only whitelist per
candidate, and a whitelist of individuals classified as a homokaryotype across
every candidate.

The homokaryotype sets are useful for asking whether a downstream signal
persists after excluding putative heterokaryotypes. Haplotype scans need further
care: `selscan`, for example, does not accept missing genotype or haplotype data.
Do not convert missing calls silently into a homozygous arrangement.

Three clusters are biologically suggestive only when the central PCA cluster
also has the expected heterozygosity and the pattern is not explained by
population, family, or batch. Two clusters may occur when one arrangement is
rare or absent, while continuous PCA scores may indicate ordinary population
structure rather than a polymorphic inversion.

The requested `cluster.k` controls the reported arrangement calls, but
`cluster_models` compares one-, two-, and three-cluster descriptions of PC1
using an approximate Gaussian BIC. `recommended_cluster_k` in the candidate
table records the lowest-BIC model. This check prevents a three-group story
from being accepted only because k-means was asked for three centres.

When `stability.replicates > 0`, regional SNPs are resampled and the PCA and
clustering are repeated. `assignment_stability` is the proportion of repeated
calls that agree after accounting for the arbitrary direction of PC1. Low
overall stability or unstable individuals near cluster boundaries should make
arrangement labels provisional.

The LD diagnostic uses the upper triangle for all individuals and the lower
triangle for the more common outer PCA cluster, treated provisionally as a
homokaryotype. This comparison is useful because mixing arrangements can create
strong LD even when LD is lower within one arrangement. It is not independent
validation: the subset was inferred from the same candidate-region genotypes.
The group-specific values are available in `candidate1$ld_summary`.

## Coverage, allele balance, and missingness artifacts

Technical summaries are generated from whatever compatible count information
was retained in the GDS, not from the original filename or input format. The
function recognizes standard `DP` and biallelic `AD` nodes and
genometranslator genotype-metadata fields named `READ_DEPTH`,
`ALLELE_REF_DEPTH`, and `ALLELE_ALT_DEPTH`. These may originate from a VCF,
DArT two-row count file, or another count-based format.

```{r technical-diagnostics, eval=FALSE}
candidate1$coverage_source
candidate1$technical_summary

candidate1$scores |>
  dplyr::select(individual, arrangement, call_rate, mean_depth,
                mean_heterozygote_allele_balance)

inv$output.files$plots$`INV-CAND-1_coverage_call_rate`
```

A depth shift among arrangement classes may reflect a real structural variant,
but it can also indicate paralogy, a copy-number variant, collapsed repeats, or
mapping bias. Allele balance far from 0.5 among called heterozygotes is another
warning. These measurements help distinguish a local genotype signal from a
coverage artifact; they do not prove which biological mechanism produced it.

## Diagnostic SNP loadings and arrangement differentiation

Regional PC1 loadings rank the SNPs contributing most strongly to the local
structure. They can guide diagnostic-panel exploration, but marker selection
and evaluation must use separate samples or resampling to avoid upward bias.

```{r loadings-differentiation, eval=FALSE}
candidate1$pc1_loadings |>
  dplyr::slice_min(loading_rank, n = 25)

candidate1$arrangement_differentiation
inv$output.files$plots$`INV-CAND-1_marker_loadings`
```

The arrangement table reports mean absolute allele-frequency differences and
pairwise Hudson FST. Hudson FST is used only as a compact description of
allele-frequency separation between inferred classes. AA, AB, and BB are not
natural populations, and the same regional SNPs helped define the classes.
Consequently, this FST is partly circular and is not independent evidence for
an inversion. Population differentiation should instead be estimated from
independently defined populations with an estimator such as
`assigner::fst_WC84()`.

Dxy is not calculated among AA, AB, and BB. In particular, AB is an inferred
heterokaryotype rather than an independently sampled evolutionary lineage.
Population-level Dxy belongs in a later comparison among genuine populations
or lineages.

## Mean imputation: what it does

Within each window, SNPs below `min.call.rate` are excluded. At each remaining
SNP, a missing genotype is temporarily replaced by the observed mean allele
dosage for that SNP. The value is used only to make covariance PCA matrices
complete; the GDS is not changed. After marker centring, an imputed
value contributes zero deviation at that SNP. It therefore supplies no evidence
that the individual carries either arrangement.

Only covariance PCA uses these imputed values. LD is calculated from observed
genotypes using pairwise-complete correlations; missing LD genotypes are not
mean-imputed.

This is intentionally simple and transparent. It prevents PCA from discarding
every individual with one missing call, but it is not a correction for biased
missingness.

## Why RADseq missingness needs special attention

RADseq missingness is often structured rather than random. Relevant causes
include:

* polymorphism at restriction sites and consequent allele dropout;
* uneven sequencing depth among samples or libraries;
* lane, plate, library-preparation, or genotyping batches;
* population differences in reference divergence or mapping quality;
* separate marker discovery or genotype calling among sample groups;
* paralogous loci and collapsed repetitive regions;
* marker density that varies with restriction sites, assembly quality, or
  filtering.

If one population or batch is missing preferentially at a set of linked SNPs,
mean imputation pulls those samples toward the local PCA centre. Depending on
the pattern, this can weaken a biological cluster, manufacture an intermediate
group, or make a batch-specific genomic region look unusual.

For every candidate, perform sensitivity analyses:

1. plot marker and individual call rate, depth, and heterozygosity;
2. cross-tabulate inferred clusters against population, plate, lane, library,
   family, and sampling date;
3. repeat the scan with stricter `min.call.rate` values;
4. repeat after excluding the most incomplete samples;
5. compare fixed-SNP and fixed-base-pair windows when marker density is uneven;
6. inspect whether the signal remains when close relatives are removed;
7. verify that marker density and assembly gaps do not define the candidate
   boundaries.

A robust biological signal should not disappear under every reasonable QC
choice, and its clusters should not be synonymous with one technical batch.

## Results folder and customised plots

Each call creates a dated `detect_inversions` folder following the standard
`radr` workflow. It contains the recorded call, TSV tables, and PNG and PDF
figures. Timestamped folders prevent a normal run from overwriting an earlier
analysis. The returned object also contains the window table, candidate table,
PCA scores, cluster summaries, LD summaries, and plot objects. A standard plot
can therefore be adjusted and saved under an additional name:

```{r save-plot, eval=FALSE}
p <- inv$output.files$plots$window_scores +
  labs(title = "Candidate inversion scan after batch QC")

ggsave(file.path(inv$path.folder, "window_scores_annotated.png"), p,
       width = 10, height = 7, dpi = 300)
```

Use `save.plots = FALSE` when only tables and returned objects are wanted. The
results folder and reproducibility tables are still created.

## Add context to genome scans

Use `genome_scan_context()` to place scan peaks beside marker density,
missingness, heterozygosity, allele frequency, LD, and candidate-region
annotations:

```{r scan-context, eval=FALSE}
context <- radr::genome_scan_context(
  data = genome,
  inversion.regions = inv,
  scan.statistics = bayescan_or_fst_table,
  window.snps = 250
)

context$context
```

For each regional peak, ask:

* Does it overlap low recombination or a centromere?
* Is marker density unusual?
* Does missingness increase in the same region?
* Is the signal restricted to one caller, mapping strategy, sequencing batch,
  or reference assembly?
* Does it remain after excluding putative inversion heterozygotes?
* Is it supported by a method based on a different genomic signature?

These questions should also guide analyses in `assigner` and `grouper`.
Population structure should be explored with the full marker set, with candidate
inversion regions excluded, and with arrangement genotypes included explicitly.
Otherwise a large haploblock can dominate clustering and make one regional
feature look like genome-wide structure.

## Interpreting coordinates

The candidate `start` and `end` values are the outer marker coordinates of the
selected windows. They depend on window size, step size, marker density,
missingness filters, and the reference assembly. Report them as an approximate
inversion-associated interval. With RADseq data, there may be a substantial gap
between the last marker outside the signal and the first marker inside it.

A defensible report might state:

> A candidate inversion-associated haploblock was detected from approximately
> X to Y Mb on LG14, supported by concordant local-PCA, regional genotype-group,
> heterozygosity, and LD patterns. These coordinates describe the marker-defined
> interval and not validated structural breakpoints.

## Interpreting the candidate evidence table

The candidate table combines continuous diagnostics with a deliberately simple
screening grade. Important columns include:

* `cluster_separation`: the smallest gap between adjacent regional-PC1 cluster
  centres, divided by the pooled within-cluster standard deviation;
* `cluster_compactness`: the proportion of regional PC1 variation explained by
  the inferred clusters;
* `recommended_cluster_k`: the one-, two-, or three-cluster PC1 model with the
  lowest approximate BIC;
* `assignment_stability`: agreement of arrangement calls across SNP-resampling
  replicates, when requested;
* `smallest_cluster_n` and `smallest_cluster_frequency`: protection against
  treating a few outlying individuals as an inversion arrangement;
* `middle_heterozygosity_excess`: middle-cluster heterozygosity minus the mean
  of the two outer clusters;
* `regional_mean_ld_r2`, `homokaryotype_mean_ld_r2`, and
  `heterokaryotype_mean_ld_r2`: LD across all individuals and within inferred
  genotype groups;
* `flanking_mean_ld_r2` and `boundary_contrast`: comparison with the immediate
  noncandidate windows;
* `internal_transition_max`: the largest score change between candidate
  windows, useful for finding complex, adjacent, or nested signals; and
* `known_region_overlap`: any user-supplied centromere, low-recombination,
  assembly-gap, or other annotation intersecting the candidate;
* `candidate_class`: a cautious interpretation such as putative
  inversion-associated haploblock, unresolved haploblock, low-recombination or
  centromeric candidate, technical or assembly-associated region, or another
  structural-variation-associated haploblock; and
* `alternative_explanations`: reminders that local PCA can reflect a
  centromere, low recombination, assembly or mapping problems, introgression,
  population-specific missingness, or another structural variant.

`three_cluster_evidence` is not merely a record that k-means was run with
`cluster.k = 3`. It additionally requires all three inferred groups to contain
at least three samples, the smallest group to represent at least 5% of samples,
and adequate separation relative to within-group spread.

`evidence_score` assigns one point for each of five observations: quantitative
three-cluster support, positive middle-cluster heterozygosity excess, positive
boundary contrast, regional LD above flanking LD, and continuity across at
least two windows. Scores of 0--2, 3--4, and 5 are labelled `weak`, `moderate`,
and `strong`, respectively. This grade prioritises candidates for inspection;
it is not a posterior probability and does not prove an inversion.

## Recommended workflow

Use the scan as one part of a staged analysis:

1. complete sample, marker, batch, depth, and missingness QC;
2. scan chromosome-specific windows for unusual local structure;
3. inspect contiguous candidates rather than isolated outlier windows;
4. evaluate regional PCA groups, heterozygosity, and LD jointly;
5. test sensitivity to filtering, window size, step size, and relatedness;
6. compare candidates with centromeres, assembly gaps, repeats, and gene
   annotations;
7. use linkage maps or haplotagging to test recombination and phase the
   alternative arrangements; and
8. seek breakpoint-spanning long reads, junction PCR, comparative assembly, or
   cytogenetic evidence before calling a structurally confirmed inversion.

For genome scans and population-structure analyses, report three deliberate
views whenever the data permit:

1. the complete genome;
2. a collinear sensitivity dataset excluding candidate inversion-associated or
   major low-recombination regions; and
3. candidate-region analyses using arrangement frequencies, heterozygosity,
   LD, differentiation, environmental associations, and homokaryotype-only
   sensitivity datasets.

Inversions should be detected, annotated, and analysed explicitly. They should
neither be ignored nor automatically removed. Use "candidate", "putative", or
"inversion-associated haploblock" until structural evidence justifies a more
definitive term.

## References

Akopyan M, Tigano A, Jacobs A, Wilder AP, Therkildsen NO (2025) Genetic
differentiation is constrained to chromosomal inversions and putative
centromeres in locally adapted populations with higher gene flow.
*Molecular Biology and Evolution*, 42, msaf092.

Campbell MA, Anderson EC, Garza JC, Pearse DE (2021) Polygenic basis and the
role of genome duplication in adaptation to similar selective environments.
*Journal of Heredity*, 112, 614-625.

Bhatia G, Patterson N, Sankararaman S, Price AL (2013) Estimating and
interpreting FST: the impact of rare variants. *Genome Research*, 23,
1514-1521.

Faria R, Johannesson K, Butlin RK, Westram AM (2019) Evolving inversions.
*Trends in Ecology & Evolution*, 34, 239-248.

Gosselin T, Anderson EC, Bradbury I (2020) assigner: assignment analysis with
GBS/RAD data using R. R package. https://doi.org/10.5281/zenodo.592677.

Hoffmann AA, Rieseberg LH (2008) Revisiting the impact of inversions in
evolution: from population genetic markers to drivers of adaptive shifts and
speciation? *Annual Review of Ecology, Evolution, and Systematics*, 39, 21-42.

Hudson RR, Slatkin M, Maddison WP (1992) Estimation of levels of gene flow from
DNA sequence data. *Genetics*, 132, 583-589.

Huang K, Andrew RL, Owens GL, Ostevik KL, Rieseberg LH (2020) Multiple
chromosomal inversions contribute to adaptive divergence of a dune sunflower
ecotype. *Molecular Ecology*, 29, 2535-2549.

Kess T, Bentzen P, Lehnert SJ, Sylvester EVA, Lien S, Kent MP,
Sinclair-Waters M, Morris CJ, Regular P, Fairweather R, Bradbury IR (2019) A
migration-associated supergene reveals loss of biocomplexity in Atlantic cod.
*Science Advances*, 5, eaav2461.

Li H, Ralph P (2019) Local PCA shows how the effect of population structure
differs along the genome. *Genetics*, 211, 289-304.

Meier JI, Salazar PA, Kučka M, Davies RW, Dréau A, Aldás I, Power OB, Nadeau NJ,
Bridle JR, Rolian C, Barton NH, McMillan WO, Jiggins CD, Chan YF (2021)
Haplotype tagging reveals parallel formation of hybrid races in two butterfly species.
*Proceedings of the National Academy of Sciences of the United States of America*,
118, e2015005118.

Mérot C (2020) Making the most of population genomic data to understand the
importance of chromosomal inversions for adaptation and speciation.
*Molecular Ecology*, 29, 2513-2516.

Pearse DE, Barson NJ, Nome T, Gao G, Campbell MA, Abadía-Cardoso A, Anderson EC,
Rundio DE, Williams TH, Naish KA, Moen T, Liu S, Kent M, Moser M, Minkley DR,
Rondeau EB, Brieuc MSO, Sandve SR, Miller MR, Cedillo L, Baruch K, Hernandez AG,
Ben-Zvi G, Shem-Tov D, Barad O, Kuzishchin K, Garza JC, Lindley ST, Koop BF,
Thorgaard GH, Palti Y, Lien S (2019) Sex-dependent dominance maintains migration
supergene in rainbow trout. *Nature Ecology & Evolution*, 3, 1731-1742.

Thompson NF, Anderson EC, Clemento AJ, Campbell MA, Pearse DE, Hearsey JW,
Kinziger AP, Garza JC (2020) A complex phenotype in salmon controlled by a
simple change in migratory timing. *Science*, 370, 609-613.

Weir BS, Cockerham CC (1984) Estimating F-statistics for the analysis of
population structure. *Evolution*, 38, 1358-1370.

Wellenreuther M, Bernatchez L (2018) Eco-evolutionary genomics of chromosomal
inversions. *Trends in Ecology & Evolution*, 33, 427-440.
[https://doi.org/10.1016/j.tree.2018.04.002](https://doi.org/10.1016/j.tree.2018.04.002)
