---
title: "Candidate copy-number variation"
author: "Thierry Gosselin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Candidate copy-number variation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## What this diagnostic does

`detect_cnv()` screens retained read depth for **candidate relative-depth
variation**. It does not establish absolute copy number, call breakpoints,
or prove a duplication. Paralogy, allele dosage and copy-number variation
are related but distinct: a duplicated locus can have constant copy number
across every individual.

McKinney et al. (2017) provide the conceptual foundation for combining
heterozygosity and allelic read-ratio evidence to investigate paralogy.
The supporting imbalance summary here is not their full HDplot statistic.
The primary candidate flag currently uses relative depth, not HDplot
classification. McKinney et al. (2018) extend this topic to allele-dosage
inference; their likelihood model is not implemented here.

Dorant et al. (2020) demonstrate the value of examining duplicated loci in
marine Rapture data. Their normalized-depth approach motivates exploration,
but radr does not reproduce that pipeline or its threshold classifications.
Jiang et al. (2018) highlight why normalization must be treated as part of
inference. The implementation here is a transparent median-ratio screen,
not TMM, CODEX2, or a validated CNV genotyper.

## Input and timing

Use a biallelic GDS retaining per-sample depth. Genotype calls alone are
insufficient. Run after sample QC, but before permanently discarding all
high-depth/paralogous loci. Previously excluded loci are not reactivated.

```{r, eval=FALSE}
cnv <- radr::detect_cnv(
  data = "study.gds",
  strata = "sample_metadata.tsv",
  batch.columns = c("PLATE", "PROJECT"),
  locus.columns = c("CHROM", "LOCUS")
)
cnv$statistics
cnv$metadata.associations
cnv$group.batch.counts
```

Locus identity must be explicit. `LOCUS` is the original read/assembly locus,
not necessarily a reference coordinate. The smallest active `VARIANT_ID`
represents each locus. Multiple SNPs sharing reads are not summed. This avoids
inflating support with SNP density, but representative-SNP depth can miss
within-locus differences. Do not substitute individual variant identifiers
for genuine locus metadata merely to make a RADseq analysis run.

## Normalization and interpretation

The geometric mean of positive depths is calculated per eligible locus.
Each sample's median ratio to these means estimates its library factor.
Scaled depths are divided by each locus's cohort median. Defaults require
100 positive-depth baseline loci per sample and 10 samples per baseline locus.
These are safeguards, not proof of adequate power or an unbiased baseline.

Missing depth remains missing; observed zero remains zero. Zero cohort-median
loci are uninformative because no relative baseline can be estimated. The
baseline assumes most loci have stable copy number. Global ploidy changes and
widespread copy-number differences can be normalized away.

At least three samples with relative depth at or above 1.5, or at or below
1/1.5, flag a candidate by default. These are **screening thresholds**, not
statistical tests. A ratio of two does not establish four genomic copies.
The output provides raw depth, relative depth, sample factors and representative
variants so each signal can be traced back to its evidence.

Allele imbalance is supporting evidence from called heterozygotes. It uses
absolute deviation from 0.5, independent of REF/ALT orientation. Missing allele
depth produces missing evidence, not a false zero imbalance.

## Technical confounding and follow-up

Metadata eta-squared is a descriptive association with log2(1 + relative depth),
not a p-value. Categories with few samples can inflate it. Review the
group-by-batch tables: if populations and plates coincide, an association does
not identify its cause. No automatic batch correction is performed.

Restriction-site mutations, capture efficiency, mapping ambiguity, DNA quality
and library preparation can mimic copy-number signals. Validate promising
loci using independent libraries, segregation, targeted assays or suitable
whole-genome/long-read evidence. WGS CNV discovery requires genomic coverage
and alignment evidence beyond a filtered SNP GDS.

## Safeguards: evidence is not the same as technical reliability

Dallaire et al. (2023) show why depth, heterozygosity and allele ratios must
be scrutinized together. A deviant locus may represent a biological duplication,
a mapping problem, or both. This diagnostic does not run ngsParalog and does
not automatically discard deviant loci from CNV discovery.

- `evidence` describes relative-depth candidates and availability of allele
  balance; it does not assign absolute copy numbers.
- `stratum.statistics` reports observed/expected heterozygosity, called sample
  counts, heterozygotes with allele depth, and imbalance separately per stratum.
  Small sample counts limit interpretation. These are descriptive summaries,
  not a calibrated deviant-SNP likelihood test.
- `technical.risks` separates supplied paralogy, repeat and mappability flags
  from observed batch associations. Unknown checks are `NOT_ASSESSED`.
- Independent validation can be recorded as `INDEPENDENT_SUPPORT`; it is never
  inferred from the depth threshold alone.
- Read-mapping audits and normalization-sensitivity benchmarking remain
  `NOT_ASSESSED`; having a risk table is not equivalent to performing them.

Supply a locus-level annotation table using exactly the same locus keys as
the GDS, with one row per locus. Flags must be logical `TRUE`, `FALSE`, or `NA`.
For example, reviewed `detect_paralogs()` results can contribute `PARALOG_FLAG`.
Aggregate SNP-level results deliberately: a negative flag requires adequate
assessment, not simply the absence of a positive SNP. Ensure assembly, sample
selection and locus definitions are appropriate; these cannot be verified
automatically from a bare flag table.

```{r, eval=FALSE}
annotations <- data.frame(
  CHROM = c("1", "1"), LOCUS = c("locus_a", "locus_b"),
  PARALOG_FLAG = c(TRUE, NA),
  LOW_MAPPABILITY = c(NA, FALSE),
  REPEAT_OVERLAP = c(TRUE, FALSE)
)
cnv <- radr::detect_cnv("study.gds", strata = "samples.tsv",
  batch.columns = "PLATE", risk.annotations = annotations)
cnv$evidence
cnv$technical.risks
cnv$stratum.statistics
```

## Why compare other methods?

Comparing approaches is good scientific practice: distinct models expose
different assumptions and failure modes. Dorant motivates the RAD/capture
application; McKinney distinguishes paralogy detection from dosage inference;
CODEX2 addresses coverage normalization; CNVpytor combines WGS depth with allele
imbalance; Genome STRiP supports population-scale multiallelic CNV inference.
Schrider and Hahn provide evolutionary context. Citing these methods does not
mean radr implements them or has completed comparative benchmarking. The
current screen should be evaluated against suitable alternatives and independent
validation before drawing biological conclusions.

## References

McKinney, G. J., Waples, R. K., Seeb, L. W., and Seeb, J. E. (2017).
Paralogs are revealed by proportion of heterozygotes and deviations in read
ratios in genotyping-by-sequencing data from natural populations.
*Molecular Ecology Resources*, 17, 656-669.
<https://doi.org/10.1111/1755-0998.12613>

McKinney, G. J., Waples, R. K., Pascal, C. E., Seeb, L. W., and Seeb, J. E.
(2018). Resolving allele dosage in duplicated loci using genotyping-by-sequencing
data: A path forward for population genetic analysis.
*Molecular Ecology Resources*, 18, 570-579.
<https://doi.org/10.1111/1755-0998.12763>

Dorant, Y., et al. (2020). Copy number variants outperform SNPs to reveal
genotype-temperature association in a marine species. *Molecular Ecology*.
<https://doi.org/10.1111/mec.15565>

Jiang, Y., et al. (2018). CODEX2: full-spectrum copy number variation detection
by high-throughput DNA sequencing. *Genome Biology*, 19, 202.
<https://doi.org/10.1186/s13059-018-1578-y>

Dallaire, X., et al. (2023). Widespread Deviant Patterns of Heterozygosity in
Whole-Genome Sequencing Due to Autopolyploidy, Repeated Elements, and Duplication.
*Genome Biology and Evolution*, 15, evad229.
<https://doi.org/10.1093/gbe/evad229>

Suvakov, M., Panda, A., Diesh, C., Holmes, I., and Abyzov, A. (2021).
CNVpytor: a tool for copy number variation detection and analysis from read depth
and allele imbalance in whole-genome sequencing. *GigaScience*, 10, giab074.
<https://doi.org/10.1093/gigascience/giab074>

Handsaker, R. E., et al. (2015). Large multiallelic copy number variations in
humans. *Nature Genetics*, 47, 296-303. <https://doi.org/10.1038/ng.3200>

Schrider, D. R., and Hahn, M. W. (2010). Gene copy-number polymorphism in nature.
*Proceedings of the Royal Society B*, 277, 3213-3221.
<https://doi.org/10.1098/rspb.2010.1180>
