---
title: "Exploring and filtering genomic data with radr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring and filtering genomic data with radr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Responsibilities of the packages

`radr` begins after genomic data have been imported and standardized.
[`genometranslator`](https://thierrygosselin.github.io/genometranslator/)
handles file formats and creates the GDS representation used throughout this
article. `radr` diagnoses the data and performs deliberate filtering. Writers
in `genometranslator` do not silently perform quality-control filtering.

The examples are not evaluated because they refer to user files.

## Start from sample metadata and an unfiltered GDS

Maintain one authoritative sample metadata table and generate the compact
strata file from it. The strata file should contain stable `INDIVIDUALS` names
and the grouping variable required for the current analysis. Consult the
[genometranslator vignette](https://thierrygosselin.github.io/genometranslator/articles/using_genometranslator.html)
for identifier, `NEW_ID`, and DArT `TARGET_ID` recommendations.

```{r import, eval=FALSE}
library(radr)

genome <- genometranslator::read_genome(
  data = "individuals.vcf.gz",
  strata = "strata.tsv"
)

genometranslator::summary_gds(genome)
```

Keep the original genomic source file untouched. A GDS is stateful: filtering
changes the active samples or markers and records those changes in the same
backing file. Assigning the result to another R object does not create an
independent copy of that GDS.

Each filtering stage creates a dated results folder. Depending on the function,
it contains the function-call arguments, diagnostic figures and tables,
blacklists, and summaries. The dated `filters_parameters_*.tsv` file records
the filters and parameter values together with before-and-after dimensions and
blacklist information. Keep these folders with the original source file and
sample metadata so that filtering decisions can be reviewed and reproduced.

## Inspect missingness before choosing what to remove

A common first question is whether to filter markers or individuals first.
There is no universal answer because each choice changes statistics calculated
for the other dimension.

`detect_ibm()` displays the raw variant-by-sample presence/absence pattern. The
goal is to reveal broad dataset structure, not to identify one labelled marker
or sample. For the first view, preserve input order: sequencing plates,
libraries, batches, and processing order can create patterns that sorting would
hide.

```{r ibm, eval=FALSE}
ibm <- detect_ibm(
  gds = genome,
  sort.individuals = "input",
  sort.markers = "input",
  filename = "initial_missingness.png"
)
```

Interpret broad patterns cautiously:

* vertical bands suggest samples with unusual missingness;
* horizontal bands suggest marker sets with unusual missingness;
* rectangular blocks can indicate batches, strata, libraries, or marker panels;
* isolated patterns can reflect technical problems, biology, or both.

Use marker- or individual-missingness sorting only as a complementary second
view. A pattern is evidence for investigation, not an automatic blacklist.

## Guided exploration

`explore_genomes()` provides an interactive overview for a new dataset and for
users learning how candidate thresholds affect it:

```{r explore, eval=FALSE}
screened <- explore_genomes(
  data = genome,
  interactive.filter = TRUE
)
```

The function creates a dated folder containing arguments, figures, helper
tables, filters, and summaries. It is opinionated and convenient, but it is not
"one function to rule them all." Marker type, sequencing design, organism,
population structure, and analysis objective determine which filters are valid
and in which order they should run.

## Build a tailored workflow

After learning the dataset, call individual functions with recorded thresholds.
The filtering order should express a hypothesis about how the data became
problematic. The two examples below deliberately use opposite orders.

### Marker noise is the dominant problem

Imagine that genotyping parameters were not tailored to the project, or that
files genotyped separately were combined after variant calling. Differences in
coverage, callers, parameter settings, reference versions, or marker discovery
can introduce a large number of inconsistently genotyped SNPs. These markers
can make otherwise acceptable samples appear to have high missingness.

Remove clear marker failures first, then recalculate individual statistics:

```{r markers-first, eval=FALSE}
markers_first <- genome |>
  filter_genotyping(
    interactive.filter = FALSE,
    filter.genotyping = 0.20
  ) |>
  filter_ma(
    interactive.filter = FALSE,
    filter.ma = 2
  ) |>
  filter_individuals(
    interactive.filter = FALSE,
    filter.individuals.missing = 0.30
  )
```

This is a downstream repair strategy. When source data are available, a
consistent joint-genotyping workflow is preferable. At minimum, verify that
combined files used compatible references, marker definitions, callers,
filters, and genotype fields. Marker filtering cannot prove that batch effects
have been removed.

### Poor samples are the dominant problem

Now imagine that the project records already identify failed or weak libraries:
some samples received few reads, had poor base quality or adapter contamination,
mapped badly, or had abnormally low coverage. Those individuals can introduce
missing calls at many markers that perform well in the rest of the project.

Remove the failed samples first, then return to `explore_genomes()` with the
individuals that remain. This resumes the normal guided discovery workflow
without assuming which marker filter should come next:

```{r samples-first, eval=FALSE}
samples_first <- genome |>
  filter_individuals(
    interactive.filter = FALSE,
    filter.individuals.missing = 0.50,
    filter.individuals.coverage.total = 1e6
  ) |>
  explore_genomes()
```

Ideally, these samples are identified before genotyping through FASTQ and
library QC: read yield, per-base quality, adapter content, duplication,
contamination, mapping rate, and depth should be reviewed with suitable tools
and project-specific expectations. Removing them in radr is a downstream
confirmation and safeguard, not a replacement for upstream QC.

The numeric thresholds in both examples are illustrations. Derive thresholds
from diagnostic figures, project design, sequencing expectations, and the
requirements of the intended analysis. When comparing alternative orders,
create separate physical copies of the original GDS file or re-import the
untouched source data. Assigning the same GDS connection to a different R
variable does not create an independent copy.

At each stage, ask what changed:

* Did filtering individuals alter marker missingness or allele counts?
* Did marker filtering change which individuals appear unusual?
* Is a signal restricted to one stratum, sequencing batch, or marker class?
* Does the chosen statistic match the intended downstream analysis?

Re-run the relevant diagnostic after a consequential filter. Avoid choosing a
threshold solely because it is conventional in another study.

## Review active filters

The parameter file documents how the dataset reached its current state. To see
which individual and marker filters are currently active in the GDS metadata,
use:

```{r list-filters, eval=FALSE}
genometranslator::list_filters(genome)
```

This current-state summary complements the chronological parameter log. Keep
both: the active-filter list answers *what is applied now*, whereas the dated
folders and parameter file document *what was done, in which order, and with
which thresholds*.

## Detection does not necessarily mean filtering

The `detect_*()` functions investigate particular structures or problems:

```{r detect, eval=FALSE}
duplicates <- detect_duplicate_genomes(data = filtered)
mixed <- detect_mixed_genomes(data = filtered)
paralogs <- detect_paralogs(data = filtered)
```

Review returned tables and figures together with sample metadata. A duplicate,
sex-linked marker, family group, or biologically divergent sample may be
important information rather than an error.

## Export only after filtering decisions

Use `genometranslator` to write the final GDS to a downstream format. The writer
does not silently improve or filter the data:

```{r export, eval=FALSE}
genometranslator::write_genome(
  data = filtered,
  output = "vcf",
  filename = "analysis_ready"
)
```

Confirm the destination method's assumptions before export. For example,
BayeScan, structure-like programs, relatedness software, and LD analyses can
require different marker, linkage, ploidy, and population treatments.

## Dependencies and external tools

Inspect requirements at any time with:

```{r dependencies, eval=FALSE}
radr_dependencies()
```

The diagnostic separates the required package foundation from optional
components. `SNPRelate` supports LD and IBS workflows; `quantreg` is used by
`sexy_markers()`; and `ragg` accelerates PNG output from `detect_ibm()`.

VCF-level filtering functions may call the external `bcftools` executable, and
`run_bayescan()` requires BayeScan. Install these through the shared Conda
`genomics` environment described in the README, then start R or RStudio from
that activated environment.

## Reproducibility checklist

For a defensible filtering workflow, retain:

* the original genomic and sample metadata files;
* the standardized strata file;
* radr argument and filter-parameter logs;
* generated blacklists and whitelists;
* diagnostic figures before and after important decisions;
* the random seed where sampling is used;
* R, radr, genometranslator, tgbase, and executable versions;
* the final GDS and the exact export command.

During active development, record the Git commit in addition to the package
version.
