---
title: "Exploring and filtering genomic data with radr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring and filtering genomic data with radr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Responsibilities of the packages

`radr` begins after genomic data have been imported and standardized.
[`genometranslator`](https://thierrygosselin.github.io/genometranslator/)
handles file formats and creates the GDS representation used throughout this
article. `radr` diagnoses the data and performs deliberate filtering. Writers
in `genometranslator` do not silently perform quality-control filtering.

The examples are not evaluated because they refer to user files.

## Start from sample metadata and an unfiltered GDS

Maintain one authoritative sample metadata table and generate the compact
strata file from it. The strata file should contain stable `INDIVIDUALS` names
and the grouping variable required for the current analysis. Consult the
[genometranslator vignette](https://thierrygosselin.github.io/genometranslator/articles/using_genometranslator.html)
for identifier, `NEW_ID`, and DArT `TARGET_ID` recommendations.

```{r import, eval=FALSE}
library(radr)

genome <- genometranslator::read_genome(
  data = "individuals.vcf.gz",
  strata = "strata.tsv"
)

genometranslator::summary_gds(genome)
```

Keep the original genomic source file untouched. A GDS is stateful: filtering
changes the active samples or markers and records those changes in the same
backing file. An open GDS object is a connection to that file, not an in-memory
copy of all its contents. Consequently, assigning the connection to another R
name does not create an independent dataset:

```{r gds-alias, eval=FALSE}
genome_copy <- genome
```

Here, `genome` and `genome_copy` refer to the same physical GDS. Filtering
through either name updates the marker or individual `FILTERS` metadata and the
active selection seen through both names. Similarly, assigning the output of a
filter to a new object does not protect the input GDS from modification.

Before continuing an analysis, inspect both the active dimensions and the
filters recorded in the metadata:

```{r inspect-gds-filters, eval=FALSE}
genometranslator::summary_gds(genome, check.sync = TRUE)
genometranslator::list_filters(genome, history = TRUE)
```

This is particularly important after an interrupted interactive analysis,
after reopening a previously used GDS, or before assuming that an object is an
unfiltered starting point.

Each filtering stage creates a dated results folder. Depending on the function,
it contains the function-call arguments, diagnostic figures and tables,
blacklists, and summaries. Filtering history is stored inside the GDS and each
dated `filters_parameters_*.tsv` is a cumulative snapshot through that stage.
It records operation order, parameter values, before-and-after dimensions, and
blacklist information. This works identically for piped and sequential calls:
the shared backing GDS, rather than the R expression, connects the history.

## Summarise the active genomic data

`summarise_genomic_data()` works directly from the active GDS selection. It
reads genotypes in chunks and reports pooled and per-stratum call rates,
genotype and allele counts, REF and ALT frequencies, minor-allele frequency,
observed and expected heterozygosity, and FIS. It does not convert the complete
dataset to a tidy genotype table and it restores the active GDS filters after
the calculation.

```{r genomic-summary, eval=FALSE}
genomic.summary <- summarise_genomic_data(
  data = genome,
  strata = "strata.tsv",
  group.column = "STRATA"
)

genomic.summary$marker.statistics
genomic.summary$stratum.statistics
genomic.summary$group.summary
```

ALT frequency and minor-allele frequency are deliberately separate. The ALT
allele is defined by the VCF record and can have a frequency greater than 0.5;
the minor allele is whichever allele is less frequent in the samples being
summarized. The stratum table also retains the pooled minor-allele identity, so
the same biological allele can be compared across groups even when it becomes
locally common or fixed. FIS is reported as missing for monomorphic markers,
where expected heterozygosity is zero and the coefficient is undefined.

## Inspect missingness before choosing what to remove

A common first question is whether to filter markers or individuals first.
There is no universal answer because each choice changes statistics calculated
for the other dimension.

`detect_ibm()` displays the raw variant-by-sample presence/absence pattern. The
goal is to reveal broad dataset structure, not to identify one labelled marker
or sample. For the first view, preserve input order: sequencing plates,
libraries, batches, and processing order can create patterns that sorting would
hide.

```{r ibm, eval=FALSE}
ibm <- detect_ibm(
  data = genome,
  sort.individuals = "input",
  sort.markers = "input",
  filename = "initial_missingness.png"
)
```

Interpret broad patterns cautiously:

* vertical bands suggest samples with unusual missingness;
* horizontal bands suggest marker sets with unusual missingness;
* rectangular blocks can indicate batches, strata, libraries, or marker panels;
* isolated patterns can reflect technical problems, biology, or both.

Use marker- or individual-missingness sorting only as a complementary second
view. A pattern is evidence for investigation, not an automatic blacklist.

## Guided exploration

`explore_genomes()` provides an interactive overview for a new dataset and for
users learning how candidate thresholds affect it:

```{r explore, eval=FALSE}
screened <- explore_genomes(
  data = genome,
  interactive.filter = TRUE
)
```

The function creates a dated folder containing arguments, figures, helper
tables, filters, and summaries. It is opinionated and convenient, but it is not
"one function to rule them all." Marker type, sequencing design, organism,
population structure, and analysis objective determine which filters are valid
and in which order they should run.

## Build a tailored workflow

After learning the dataset, call individual functions with recorded thresholds.
The filtering order should express a hypothesis about how the data became
problematic. The two examples below deliberately use opposite orders.

### Computational strategy: remove clear failures before expensive comparisons

Runtime and memory depend primarily on the number of active markers, number of
individuals, availability of depth fields, requested figures, and number of
cores. The order therefore matters computationally as well as biologically. A
filter applied early to 400,000 markers can make every subsequent genotype,
coverage, LD, distance, and local-PCA calculation substantially cheaper.

The following timing classes are intentionally approximate for contemporary
desktop hardware. Tens of thousands of markers and hundreds of individuals may
finish near the lower end; several hundred thousand markers and 1,000-2,000
individuals can move a step into the next class.

| Operation | Typical scale | Main cost and strategy |
|---|---:|---|
| Metadata validation, existing VCF filters, reproducibility fields, and monomorphic checks | seconds to a few minutes | Usually inexpensive; run first when applicable. |
| `filter_genotyping()` | seconds to a few minutes | Very fast even for large datasets and often removes many markers. It is usually the first marker-targeting filter to run. |
| `filter_ma()` | seconds to several minutes | Low memory cost with substantial potential to remove low-information markers. It is usually the second marker-targeting filter. Prefer MAC when preservation of the allele-frequency spectrum matters. |
| `filter_snp_number()` | seconds to a few minutes | Fast, but usually removes fewer markers. Use only when `LOCUS` represents a genuine shared read or RAD locus; it is generally not informative for one-record-per-locus reference VCFs. |
| `filter_coverage()` | minutes to tens of minutes | Can remove a substantial number of markers, but reading and summarizing DP/AD is computationally and memory intensive. Run it last among the routine marker-targeting filters, after the dataset has already been reduced. |
| Individual missingness, heterozygosity, and coverage | minutes to tens of minutes | Recalculate after consequential marker filtering; remove known failed libraries earlier when project QC already identifies them. |
| Short-distance LD | seconds to minutes | Cheap only when genuine shared RAD/read `LOCUS` identifiers exist; it is skipped for one-record-per-locus reference VCFs. |
| Long-distance LD | minutes to hours | One of the expensive filters; cost grows with active marker count, chromosome size, missingness-aware selection, and requested figures. Run after basic marker and sample QC. |
| Pairwise genome comparisons, duplicate/mixed-genome diagnostics, HWE by groups, local PCA, and inversion diagnostics | minutes to hours | Potentially expensive in markers, samples, groups, windows, or pairs. Reserve for the reduced dataset unless the scientific question requires unfiltered data. |

A useful default computational order for the routine marker filters is:

1. Validate samples, caller provenance, metadata, and already assigned VCF
   filters. Remove samples known from laboratory or sequencing QC to have
   failed completely.
2. Run `filter_genotyping()`. It is quick, works well on very large datasets,
   and frequently provides the largest early reduction in marker number.
3. Run `filter_ma()` with a defensible global MAC threshold. It has a relatively
   small memory footprint and can remove many low-information markers. Low-count alleles contribute little
   information to many analyses and can make later calculations expensive or
   unstable. Preserve the unfiltered master when the downstream analysis needs
   the rare-allele spectrum.
4. Run `filter_snp_number()` when loci genuinely group several SNP records.
   This filter is fast, although it commonly removes fewer markers than the two
   preceding filters. Skip it when every VCF record has a unique `LOCUS`, as is
   common for reference-based FreeBayes data.
5. Run `filter_coverage()` last among the routine marker-targeting filters.
   Coverage and allele-depth information can identify many technical failures,
   but DP/AD extraction and summarization are substantially more expensive than
   the preceding filters.
6. Recalculate individual statistics and remove supported sample failures.
7. Run short-distance pruning only when loci genuinely group several SNP
   records. Run long-distance LD and the more expensive pairwise or windowed
   diagnostics after the dataset has been reduced.

In compact form, the recommended marker-filtering sequence is therefore:

```{r efficient-marker-pipeline, eval=FALSE}
filtered <- genome |>
  radr::filter_genotyping() |>
  radr::filter_ma() |>
  radr::filter_snp_number() |> # omit when LOCUS does not group genuine reads
  radr::filter_coverage()
```

This is not a universal biological filter order. For example, severely failed
samples can inflate marker missingness and should be removed first, whereas
widespread bad markers can make good samples appear poor. Use the initial
diagnostics to decide which problem dominates, record before-and-after
dimensions and elapsed time, and recalculate statistics affected by each major
decision. A fast filter is not automatically a valid filter, and computational
convenience should not erase variation required by the intended analysis.

### Chain filters with a pipe

Most radr filters take `data` as their first argument, update the active GDS,
and return its connection. They can therefore be chained directly with the
native R pipe:

```{r pipe-filters, eval=FALSE}
filtered <- genome |>
  radr::filter_genotyping() |>
  radr::filter_ma()
```

The result returned by `filter_genotyping()` becomes the first argument of
`filter_ma()`. No placeholder is required. This works in interactive mode too:
the first filter completes its questions and updates the GDS before the second
filter begins.

For a reproducible scripted analysis, provide the thresholds explicitly:

```{r pipe-scripted, eval=FALSE}
filtered <- genome |>
  radr::filter_genotyping(
    interactive.filter = FALSE,
    filter.genotyping = 0.20
  ) |>
  radr::filter_ma(
    interactive.filter = FALSE,
    filter.ma = 2
  )
```

If an explicit placeholder is useful, the native `|>` pipe uses `_`, not `.`:

```{r pipe-placeholder, eval=FALSE}
filtered <- radr::filter_genotyping(data = genome) |>
  radr::filter_ma(data = _)
```

The `.` placeholder belongs to the magrittr pipe `%>%`:

```{r magrittr-placeholder, eval=FALSE}
filtered <- radr::filter_genotyping(data = genome) %>%
  radr::filter_ma(data = .)
```

Prefer the first form when `data` is the first argument. Remember that piping
does not create a new physical GDS: each filter updates the same backing file,
then passes its connection to the next function. In the example above,
`filtered` and `genome` remain connections to that same modified file; the new
object name is not a snapshot of the result.

To compare alternative thresholds or filter orders, create separate physical
GDS files from an unfiltered master *before* running either workflow. Close its
open connection before copying the file:

```{r independent-gds-workflows, eval=FALSE}
# Create and preserve this master before applying any radr filter.
master_path <- "genome_unfiltered.gds"

SeqArray::seqClose(genome)

stopifnot(file.copy(master_path, "genome_workflow_A.gds"))
stopifnot(file.copy(master_path, "genome_workflow_B.gds"))

genome_A <- genometranslator::read_genome("genome_workflow_A.gds")
genome_B <- genometranslator::read_genome("genome_workflow_B.gds")

genometranslator::list_filters(genome_A)
genometranslator::list_filters(genome_B)
```

Copying a GDS after filters have already been applied also copies their recorded
state. If no clean master exists, recreate one from the original VCF, DArT, or
other source file. `genometranslator::reset_filters()` can deliberately
reactivate selected or all markers and individuals, but it modifies the same
GDS and is not a substitute for independent files when comparing workflows.

### What "SNPs per locus" means for STACKS and FreeBayes

The interpretation of `filter_snp_number()` depends on whether the input
preserves a genuine RAD/read locus identifier. STACKS reconstructs RAD loci and
associates multiple SNPs with the same locus. With that representation, asking
whether a roughly 70-100-bp sequence contains three or more SNPs can be informative.
An unusually dense locus may reflect paralogy, poor locus assembly,
misalignment, unexpectedly high local diversity, or unsuitable assembly
parameters. The appropriate threshold still depends on read length, organism,
sequencing design, and expected diversity.

FreeBayes uses the word *haplotype* differently. It is haplotype-aware while
evaluating nearby alleles and can emit a SNP, multi-nucleotide polymorphism,
indel, or complex allele as a single VCF record. A FreeBayes record is therefore
not automatically a reconstructed RAD locus. In particular, the length of its
REF or ALT allele is not the number of SNPs observed on one read.

For a reference-based VCF, `filter_snp_number()` counts VCF records that share
the stored `LOCUS`. If the upstream workflow did not retain a common RAD/read
locus identifier, each VCF record has a unique locus. The original read
boundaries cannot be recovered unambiguously from genomic coordinates alone,
so radr reports one SNP per locus and skips this filter. Short-distance pruning
in `filter_ld()` is skipped for the same reason, while genomic long-distance LD
pruning remains available.

For FreeBayes data without retained RAD/read grouping, place greater emphasis
on genotype quality, allele balance, depth and coverage, excess
heterozygosity, mapping diagnostics, and genomic LD. Variant density can also
be summarized in windows such as 70, 100, or 150 bp, but such windows are
genomic diagnostics: they should not be described automatically as reads or
RAD loci. If SNP-per-read filtering is required, preserve a shared locus or
fragment identifier upstream and carry it into the GDS metadata.

### Marker noise is the dominant problem

Imagine that genotyping parameters were not tailored to the project, or that
files genotyped separately were combined after variant calling. Differences in
coverage, callers, parameter settings, reference versions, or marker discovery
can introduce a large number of inconsistently genotyped SNPs. These markers
can make otherwise acceptable samples appear to have high missingness.

Remove clear marker failures first, then recalculate individual statistics:

```{r markers-first, eval=FALSE}
markers_first <- genome |>
  filter_genotyping(
    interactive.filter = FALSE,
    filter.genotyping = 0.20
  ) |>
  filter_ma(
    interactive.filter = FALSE,
    filter.ma = 2
  ) |>
  filter_individuals(
    interactive.filter = FALSE,
    filter.individuals.missing = 0.30
  )
```

This is a downstream repair strategy. When source data are available, a
consistent joint-genotyping workflow is preferable. At minimum, verify that
combined files used compatible references, marker definitions, callers,
filters, and genotype fields. Marker filtering cannot prove that batch effects
have been removed.

### Poor samples are the dominant problem

Now imagine that the project records already identify failed or weak libraries:
some samples received few reads, had poor base quality or adapter contamination,
mapped badly, or had abnormally low coverage. Those individuals can introduce
missing calls at many markers that perform well in the rest of the project.

Remove the failed samples first, then return to `explore_genomes()` with the
individuals that remain. This resumes the normal guided discovery workflow
without assuming which marker filter should come next:

```{r samples-first, eval=FALSE}
samples_first <- genome |>
  filter_individuals(
    interactive.filter = FALSE,
    filter.individuals.missing = 0.50,
    filter.individuals.coverage.total = 1e6
  ) |>
  explore_genomes()
```

Ideally, these samples are identified before genotyping through FASTQ and
library QC: read yield, per-base quality, adapter content, duplication,
contamination, mapping rate, and depth should be reviewed with suitable tools
and project-specific expectations. Removing them in radr is a downstream
confirmation and safeguard, not a replacement for upstream QC.

The numeric thresholds in both examples are illustrations. Derive thresholds
from diagnostic figures, project design, sequencing expectations, and the
requirements of the intended analysis. When comparing alternative orders,
create separate physical copies of the original GDS file or re-import the
untouched source data. Assigning the same GDS connection to a different R
variable does not create an independent copy.

At each stage, ask what changed:

* Did filtering individuals alter marker missingness or allele counts?
* Did marker filtering change which individuals appear unusual?
* Is a signal restricted to one stratum, sequencing batch, or marker class?
* Does the chosen statistic match the intended downstream analysis?

Re-run the relevant diagnostic after a consequential filter. Avoid choosing a
threshold solely because it is conventional in another study.

## Review active filters

The parameter file documents how the dataset reached its current state. To see
which individual and marker filters are currently active in the GDS metadata,
use:

```{r list-filters, eval=FALSE}
genometranslator::list_filters(genome)

# Include the chronological history and export a cumulative snapshot
filters <- genometranslator::list_filters(
  genome,
  history = TRUE,
  filename = "filters_parameters_complete.tsv"
)
```

The active-filter list answers *what is applied now*. `filters$history` answers
*what was done, in which order, and with which thresholds*, including an
operation that removed no markers or a filter that was later reset.

Older GDS files do not yet contain this embedded history. Their existing
parameter files can be imported once, in chronological order, using either the
files themselves or their result folders:

```{r import-filter-history, eval=FALSE}
genometranslator::import_filter_history(
  genome,
  paths = c(
    "06_filter_genotyping",
    "07_filter_ma"
  )
)
```

### Identifier-level filtering audit

For GDS workflows, every radr operation using the common filtering mechanism
also writes a complete identifier-level audit in its results folder. This is
not limited to functions named `filter_*`: state-changing branches of
`detect_duplicate_genomes()` and `detect_mixed_genomes()` are audited in the
same way.

The `filter_audit_manifest.tsv` file gives, for each operation, the number of
active markers and individuals before filtering, newly removed, and kept. It
points to four corresponding tables:

- `audit_*_markers_removed.tsv` and `audit_*_markers_kept.tsv`;
- `audit_*_individuals_removed.tsv` and
  `audit_*_individuals_kept.tsv`.

Empty removal tables are intentional: they prove that a completed operation
did not newly remove that entity type. The compact chronological history stays
inside the GDS, while these potentially large tables remain in the operation
folder. Diagnostic functions that only *suggest* a blacklist without changing
the GDS are not recorded as applied filters. Likewise, VCF-to-VCF utilities
audit their external output files rather than the GDS, and genotype-masking
functions alter calls rather than the marker or individual filter state.

## Detection does not necessarily mean filtering

The `detect_*()` functions investigate particular structures or problems:

```{r detect, eval=FALSE}
duplicates <- detect_duplicate_genomes(data = filtered)
mixed <- detect_mixed_genomes(data = filtered)
paralogs <- detect_paralogs(data = filtered)
rare.alleles <- detect_rare_alleles(data = filtered)
microsatellites <- detect_microsatellites(data = filtered)
```

Review returned tables and figures together with sample metadata. A duplicate,
sex-linked marker, family group, or biologically divergent sample may be
important information rather than an error.

`detect_paralogs()` implements a conservative HDplot diagnostic using
heterozygosity among called samples and REF/ALT read-ratio deviation among
heterozygous genotypes. It reports pooled and per-stratum evidence, MAF,
missingness, evidence sufficiency, and transparent candidate categories.
Candidate labels are not proof of paralogy: population structure, sex linkage,
copy-number variation, collapsed repeats, contamination, and mapping or
genotyping errors can produce similar signals. Inspect
`paralogs$strata.statistics` and validate important candidates with sequence,
reference-placement, segregation, or copy-number evidence before removal.

`detect_rare_alleles()` recalculates the observed minor allele, carrier
count, supporting read depth, and heterozygote balance from the active GDS. It
does not assume ALT is minor and does not treat weak evidence as proof of null
alleles or allele dropout. Pooled and per-stratum warnings identify markers for
review; filtering remains a separate, auditable decision.

`detect_microsatellites()` scans marker sequences stored in the GDS for
perfect tandem repeats. Its motif lengths and minimum repeat counts are
explicit parameters saved with the results. A sequence motif is a candidate
microsatellite, not evidence that the locus is polymorphic. Markers lacking a
sequence are reported separately and are not classified as repeat-free.

### Population-genetic diagnostics and filters

The population-genetic utilities operate directly on the active GDS selection.
They calculate statistics from the current samples and markers, write explicit
evidence tables, and restore temporary selections. The three filters below
then record permanent marker or individual decisions in the standard audit:

```{r population-diagnostics, eval=FALSE}
biallelic <- detect_biallelic_problems(filtered)
het.outliers <- detect_het_outliers(filtered, strata = sample.metadata)
diversity <- pi(filtered, strata = sample.metadata)
fh <- ibdg_fh(filtered, strata = sample.metadata)
private <- private_alleles(filtered, strata = sample.metadata)

filtered <- filtered |>
  filter_fis(strata = sample.metadata, fis.min.threshold = -0.5,
             fis.max.threshold = 0.5) |>
  filter_het(strata = sample.metadata, marker.range = c(0, 0.8)) |>
  filter_hwe(strata = sample.metadata, p.threshold = 1e-4,
             adjustment = "BH")
```

`detect_biallelic_problems()` distinguishes the alleles declared in each VCF
record from alleles observed in active samples. A declared multiallelic record
does not become a biallelic record merely because only two alleles remain in
the current sample subset.

`pi()` reports finite-sample-corrected nucleotide diversity. Its default is
mean diversity per callable variant; per-base diversity requires an explicit
callable-sequence denominator. Individual heterozygosity is returned under its
own name and is not presented as individual nucleotide diversity.

`ibdg_fh()` calculates the individual homozygosity coefficient on a consistent
count scale. It reports expectations from both stratum-specific and pooled
allele frequencies, so population structure is visible rather than silently
absorbed into one estimate.

Private alleles are alleles observed in exactly one stratum, subject to allele
count and call-rate controls. Private haplotypes are stricter:
`private_haplotypes()` requires multiple variants sharing a genuine `LOCUS`
and phased genotypes. It does not rename an unphased per-marker genotype as a
haplotype.

HWE, FIS, and heterozygosity filters are quality-control tools, not automatic
tests of biological validity. Population mixture, family structure, selection,
sex linkage, inversions, and technical errors can all cause departures. Use
strata deliberately, inspect the evidence tables, and compare sensitivity to
reasonable thresholds before removing markers.

### Screen for candidate sex-linked markers

`sexy_markers()` accepts only a GDS filepath or an open GDS object. It respects
the active marker and sample selections, reads the GDS in chunks, and restores
the selections when it finishes. It does not run filters or alter the GDS.
See `vignette("detecting_sex_linked_markers", package = "radr")` for the
biological background, statistical interpretation, alternative methods, and
validation workflow.

```{r sex-markers, eval=FALSE}
sex.markers <- sexy_markers(
  data = filtered,
  strata = sample.metadata,
  sex.column = "SEX",
  require.significance = TRUE
)

sex.markers$candidates
```

The screen compares three complementary signals between known females and
males:

- marker presence identifies Y-like and W-like candidates;
- heterozygosity identifies X-like and Z-like candidates;
- normalized read depth provides a second X-like or Z-like signal when depth
  is available in the GDS.

The complete `sex_marker_statistics.tsv` file reports effect sizes, p-values,
and FDR values for every active marker. The smaller
`candidate_sex_markers.tsv` file applies the thresholds recorded in the
argument file. Candidate labels describe the direction of a signal; they do
not prove that a marker is physically located on a sex chromosome.

For classification, `sex_assignment_panel.tsv` contains only assignment-ready
Y-like and W-like presence/absence candidates. Validate this panel in samples
that were not used for marker discovery, then pass it to
`assigner::assign_genetic_sex()`. The assignment step belongs in `assigner`
because discovery accuracy, individual classification, and validation are
separate inferential tasks. Recorded sex is retained for comparison and is
never overwritten.

Include plate, lane, library, population, and other relevant columns in
`strata`. The `sex_metadata_audit.tsv` file measures their association with
recorded sex. Strong confounding can make technical dropout look sex-linked.
Within-sample depth normalization reduces global sequencing-depth differences,
but it cannot remove marker-specific batch effects. Confirm important markers
in independent samples and, where possible, with genome placement or targeted
validation.

## Export only after filtering decisions

Use `genometranslator` to write the final GDS to a downstream format. The writer
does not silently improve or filter the data:

```{r export, eval=FALSE}
genometranslator::write_genome(
  data = filtered,
  output = "vcf",
  filename = "analysis_ready"
)
```

Confirm the destination method's assumptions before export. For example,
BayeScan, structure-like programs, relatedness software, and LD analyses can
require different marker, linkage, ploidy, and population treatments.

## Dependencies and external tools

Inspect requirements at any time with:

```{r dependencies, eval=FALSE}
radr_dependencies()
```

The diagnostic separates the required package foundation from optional
components. `SNPRelate` supports LD and IBS workflows, and `ragg` accelerates
PNG output from `detect_ibm()`.

VCF-level filtering functions may call the external `bcftools` executable, and
`run_bayescan()` requires BayeScan. Install these through the shared Conda
`genomics` environment described in the README, then start R or RStudio from
that activated environment.

## Reproducibility checklist

For a defensible filtering workflow, retain:

* the original genomic and sample metadata files;
* the standardized strata file;
* radr argument and filter-parameter logs;
* generated blacklists and whitelists;
* diagnostic figures before and after important decisions;
* the random seed where sampling is used;
* R, radr, genometranslator, tgbase, and executable versions;
* the final GDS and the exact export command.

During active development, record the Git commit in addition to the package
version.
